---
title: "Untitled"
output: html_document
date: "2025-05-28"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = F, warning = F)
```


```{r}
library(tidyverse)
library(survey)
library(janitor)
library(sf)
library(innovar)
library(fastDummies)
library(ggcorrplot)
library(corrplot)
library(factoextra)
library(spdep)
library(INLA)
library(car)
library(data.table)
```

This R script has two parts:
I. Obtain direct estimations for wealth index and education level — including their recategorized versions (edu: at least completed primary education; wi: at least in the 3rd wi level).
II. Merge the direct estimations with the PCA dataset and the full census dataset (2010–2020). The resulting datasets are saved to be used in 02_SAE_INLA_II_part.R.

# Dataset from Loreto districts
```{r}
data("Peru")
distritos<- Peru %>% mutate(ubigeo = as.character(ubigeo)) %>% 
  filter(dep == "LORETO")
```

# 1. data ENDES from 2010-2020
```{r}
wi_edu_2010_2020 <-read.csv("./data/wi_edu_2010_2020.csv") %>% mutate(ubigeo = as.character(ubigeo))
```

# 2. data censo 
```{r}
## total census data from 2017 (Loreto districts)
censo_final_loreto <- read.csv("./data/censo_2017/01.censo_final_loreto_2017.csv") %>% mutate(ubigeo = as.character(ubigeo))
```

```{r}
## Data obtained through PCA
censo_pca_20 <- read.csv("./data/censo_2017/censo_2017_pca_20.csv") %>% mutate(ubigeo = as.character(ubigeo))
```

# 3. Direct estimations with ENDES

```{r}
data_final_nest<-
  wi_edu_2010_2020 %>% 
  group_by(year) %>% 
  nest() %>% 
  mutate(
    svydesign_endes= map(.x=data,
                         .f = ~svydesign(ids = ~hv001, strata = ~hv022, 
                                         weights = ~hv005, data = .x, nest = T)),
    
    wi_directa = map(.x = svydesign_endes,
                     .f = ~svyby(~hv270, ~ubigeo, design = .x, svymean, keep.var = T)),
    
    edu_directa = map(.x = svydesign_endes,
                      .f = ~svyby(~hv109, ~ubigeo, design = .x, svymean, keep.var = T)),
    
    
    
    wi_directa_recat = map(.x = svydesign_endes,
                           .f = ~svyby(~hv270_recat, ~ubigeo, design = .x, svymean, keep.var = T)),
    
    
    edu_directa_recat = map(.x = svydesign_endes,
                            .f = ~svyby(~hv109_recat, ~ubigeo, design = .x, svymean, keep.var = T))
  )
```


```{r}
## EDA-education
df_long<-
  data_final_nest %>% 
  select(year,edu_directa_recat) %>% 
  mutate(
    df = map(.x = edu_directa_recat,
                  .f = ~.x %>% pivot_longer(
                    cols = starts_with("hv109"), 
                    names_to = "categoria", 
                    values_to = "estimacion"
                  ))
  )
```


```{r}
df_long %>% 
  select(year,df) %>% 
  unnest(df) %>% 
ggplot(aes(x = estimacion)) +
  geom_histogram(binwidth = 0.05, fill = "lightblue", color = "black", alpha = 0.7) +
 facet_grid(categoria~year) +
  theme_minimal() +
  labs(title = "Distribución de las estimaciones directas por categoría", x = "Estimación Directa", y = "Frecuencia")
```



# 4. Merging PCA and full census datasets with direct estimations
Since the datasets may contain values of 0 and 1, and we will transform proportions to the logit form,
we applied a correction: if the direct estimation is equal to 0, we add a small epsilon value. If the proportion is equal to 1, 
we subtract an epsilon value.
```{r}
epsilon <- 0.005 


direct_estimation_censo_total <- 
  data_final_nest %>%
  select(year,edu_directa,wi_directa,edu_directa_recat,wi_directa_recat) %>% 
  mutate(
    
    edu_censo_cpa = map(.x = edu_directa,
                    .f = ~distritos %>% 
                      full_join(
                        censo_pca_20 %>% 
                          left_join(.x) %>% 
                          
                          mutate(
                            across(.cols = hv109NE:hv109SuE, .f = ~.x * (1 - 2 * epsilon) + epsilon, .names = '{.col}_cor'),
                            
                            across(.cols = hv109NE_cor:hv109SuE_cor, .f =~logit(.x), .names = '{.col}_logit'),
                            
                            se_hv109NE_cor = (se.hv109NE^2)/(hv109NE_cor^2 * (1-hv109NE_cor)^2),
                            se_hv109PE_cor = (se.hv109PE^2)/(hv109PE_cor^2 * (1-hv109PE_cor)^2),
                            se_hv109SE_cor = (se.hv109SE^2)/(hv109SE_cor^2 * (1-hv109SE_cor)^2),
                            se_hv109SuE_cor = (se.hv109SuE^2)/(hv109SuE_cor^2 * (1-hv109SuE_cor)^2)
                          ) %>% 
                          
                          mutate(
                            across(se_hv109NE_cor:se_hv109SuE_cor, .f=~ifelse(.x <1e-6,1e-6,.x)) # similar to epsilon for DE
                          )
                      
                      ) %>% 
                      group_by(ubigeo) %>% 
                      mutate(
                        id.sp = cur_group_id()
                      ) %>% 
                      ungroup()),
    
    
    
    
    wi_censo_cpa= map(.x = wi_directa,
                   .f = ~distritos %>% 
                     full_join(
                       censo_pca_20 %>% 
                         left_join(.x) %>% 
                         
                         mutate(
                           across(.cols = hv2701st:hv2705th, .f = ~.x * (1 - 2 * epsilon) + epsilon, .names = '{.col}_cor'),
                           
                           across(.cols = hv2701st_cor:hv2705th_cor, .f =~logit(.x), .names = '{.col}_logit'),
                           
                           se_hv2701st_cor = (se.hv2701st^2)/(hv2701st_cor^2 * (1-hv2701st_cor)^2),
                           se_hv2702nd_cor = (se.hv2702nd^2)/(hv2702nd_cor^2 * (1-hv2702nd_cor)^2),
                           se_hv2703rd_cor = (se.hv2703rd^2)/(hv2703rd_cor^2 * (1-hv2703rd_cor)^2),
                           se_hv2704th_cor = (se.hv2704th^2)/(hv2704th_cor^2 * (1-hv2704th_cor)^2),
                           se_hv2705th_cor = (se.hv2705th^2)/(hv2705th_cor^2 * (1-hv2705th_cor)^2)
                         ) %>% 
                         
                         mutate(
                           across(se_hv2701st_cor:se_hv2705th_cor, .f=~ifelse(.x <1e-6,1e-6,.x))
                         )
                       
                     ) %>% 
                     group_by(ubigeo) %>% 
                     mutate(
                       id.sp = cur_group_id()
                     ) %>% 
                     ungroup()),
    
    
    edu_censo_cpa_recat = map(.x = edu_directa_recat,
                        .f = ~distritos %>% 
                          full_join(
                            censo_pca_20 %>% 
                              left_join(.x) %>% 
                              
                              mutate(
                                across(.cols = hv109_recat, .f = ~.x * (1 - 2 * epsilon) + epsilon, .names = '{.col}_cor'),
                                
                                across(.cols = hv109_recat_cor, .f =~logit(.x), .names = '{.col}_logit'),
                                
                                se_hv109_recat_cor = (se^2)/(hv109_recat_cor^2 * (1-hv109_recat_cor)^2)
                              ) %>% 
                              
                              mutate(
                                across(se_hv109_recat_cor, .f=~ifelse(.x <1e-6,1e-6,.x))
                              )
                            
                          ) %>% 
                          group_by(ubigeo) %>% 
                          mutate(
                            id.sp = cur_group_id()
                          ) %>% 
                          ungroup()),
    
    
    wi_censo_cpa_recat= map(.x = wi_directa_recat,
                      .f = ~distritos %>% 
                        full_join(
                          censo_pca_20 %>% 
                            left_join(.x) %>% 
                            
                            mutate(
                              across(.cols = hv270_recat, .f = ~.x * (1 - 2 * epsilon) + epsilon, .names = '{.col}_cor'),
                              
                              across(.cols = hv270_recat_cor, .f =~logit(.x), .names = '{.col}_logit'),
                              
                              se_hv270_recat_cor = (se^2)/(hv270_recat_cor^2 * (1-hv270_recat_cor)^2)
                            ) %>% 
                            
                            mutate(
                              across(se_hv270_recat_cor, .f=~ifelse(.x <1e-6,1e-6,.x))
                            )
                          
                        ) %>% 
                        group_by(ubigeo) %>% 
                        mutate(
                          id.sp = cur_group_id()
                        ) %>% 
                        ungroup()),
    
    
    edu_censo_total = map(.x = edu_directa,
                          .f = ~distritos %>% 
                            full_join(
                              censo_final_loreto %>%
                                left_join(.x) %>% 
                                
                                mutate(
                                  across(.cols = hv109NE:hv109SuE, .f = ~.x * (1 - 2 * epsilon) + epsilon, .names = '{.col}_cor'),
                                  
                                  across(.cols = hv109NE_cor:hv109SuE_cor, .f =~logit(.x), .names = '{.col}_logit'),
                                  
                                  se_hv109NE_cor = (se.hv109NE^2)/(hv109NE_cor^2 * (1-hv109NE_cor)^2),
                                  se_hv109PE_cor = (se.hv109PE^2)/(hv109PE_cor^2 * (1-hv109PE_cor)^2),
                                  se_hv109SE_cor = (se.hv109SE^2)/(hv109SE_cor^2 * (1-hv109SE_cor)^2),
                                  se_hv109SuE_cor = (se.hv109SuE^2)/(hv109SuE_cor^2 * (1-hv109SuE_cor)^2)
                                ) %>% 
                                
                                mutate(
                                  across(se_hv109NE_cor:se_hv109SuE_cor, .f=~ifelse(.x <1e-6,1e-6,.x))
                                )
                              
                            ) %>% 
                            group_by(ubigeo) %>% 
                            mutate(
                              id.sp = cur_group_id()
                            ) %>% 
                            ungroup()),
    
    
    
    
    wi_censo_total = map(.x = wi_directa,
                         .f = ~distritos %>% 
                           full_join(
                             censo_final_loreto %>% 
                               left_join(.x) %>% 
                               
                               mutate(
                                 across(.cols = hv2701st:hv2705th, .f = ~.x * (1 - 2 * epsilon) + epsilon, .names = '{.col}_cor'),
                                 
                                 across(.cols = hv2701st_cor:hv2705th_cor, .f =~logit(.x), .names = '{.col}_logit'),
                                 
                                 se_hv2701st_cor = (se.hv2701st^2)/(hv2701st_cor^2 * (1-hv2701st_cor)^2),
                                 se_hv2702nd_cor = (se.hv2702nd^2)/(hv2702nd_cor^2 * (1-hv2702nd_cor)^2),
                                 se_hv2703rd_cor = (se.hv2703rd^2)/(hv2703rd_cor^2 * (1-hv2703rd_cor)^2),
                                 se_hv2704th_cor = (se.hv2704th^2)/(hv2704th_cor^2 * (1-hv2704th_cor)^2),
                                 se_hv2705th_cor = (se.hv2705th^2)/(hv2705th_cor^2 * (1-hv2705th_cor)^2)
                               ) %>% 
                               
                               mutate(
                                 across(se_hv2701st_cor:se_hv2705th_cor, .f=~ifelse(.x <1e-6,1e-6,.x))
                               )
                             
                           ) %>% 
                           group_by(ubigeo) %>% 
                           mutate(
                             id.sp = cur_group_id()
                           ) %>% 
                           ungroup())
  )
```


# 5. save datasets
```{r}
edu_censo_final=direct_estimation_censo_total %>% select(year,edu_censo_cpa) %>% 
  unnest(edu_censo_cpa) %>% ungroup() %>% select(-geometry)

wi_censo_final=direct_estimation_censo_total %>% select(year,wi_censo_cpa) %>% 
  unnest(wi_censo_cpa) %>% ungroup() %>% select(-geometry)

edu_censo_final_recat=direct_estimation_censo_total %>% select(year,edu_censo_cpa_recat) %>%
  unnest(edu_censo_cpa_recat) %>% ungroup() %>% select(-geometry)

wi_censo_final_recat=direct_estimation_censo_total %>% select(year,wi_censo_cpa_recat) %>% 
  unnest(wi_censo_cpa_recat) %>% ungroup() %>% select(-geometry)

#edu_censo_total=direct_estimation_censo_total %>% select(year,edu_censo_total) %>% unnest(edu_censo_total) %>% ungroup() %>% select(-geometry)
#wi_censo_total=direct_estimation_censo_total %>% select(year,wi_censo_total) %>% unnest(wi_censo_total) %>% ungroup() %>% select(-geometry)
```


```{r}
write.csv(edu_censo_final, "./data/edu_censo_final.csv", row.names = F)
write.csv(wi_censo_final, "./data/wi_censo_final.csv", row.names = F)

write.csv(edu_censo_final_recat, "./data/edu_censo_final_recat.csv", row.names = F)
write.csv(wi_censo_final_recat, "./data/wi_censo_final_recat.csv", row.names = F)

#write.csv(edu_censo_total, "./data/edu_censo_total.csv", row.names = F)
#write.csv(wi_censo_total, "./data/wi_censo_total.csv", row.names = F)

```

