---
title: "Untitled"
output: html_document
date: "2025-05-28"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = F, warning = F)
```


```{r}
library(tidyverse)
library(survey)
library(janitor)
library(sf)
library(innovar)
library(fastDummies)
library(ggcorrplot)
library(corrplot)
library(factoextra)
library(spdep)
library(INLA)
library(car)
library(data.table)
```

This R script has two parts:
I. Obtain direct estimations for wealth index and education level — including their recategorized versions (edu: at least completed primary education; wi: at least in the 3rd wi level).
II. Merge the direct estimations with the PCA dataset and the full census dataset (2010–2020). The resulting datasets are saved to be used in 02_SAE_INLA_II_part.R.

# Dataset from Loreto districts
```{r}
data("Peru")
distritos<- Peru %>% mutate(ubigeo = as.character(ubigeo)) %>% 
  filter(dep == "LORETO")
```

# 1. data ENDES from 2010-2020
```{r}
wi_edu_2010_2020 <-read.csv("./data/wi_edu_2010_2020.csv") %>% mutate(ubigeo = as.character(ubigeo))
```

# 2. data censo 
```{r}
## total census data from 2017 (Loreto districts)
censo_final_loreto <- read.csv("./data/censo_2017/01.censo_final_loreto_2017.csv") %>% mutate(ubigeo = as.character(ubigeo))
```

```{r}
## Data obtained through PCA
censo_pca_20 <- read.csv("./data/censo_2017/censo_2017_pca_20.csv") %>% mutate(ubigeo = as.character(ubigeo))
```

# 3. Direct estimations with ENDES

```{r}

options(survey.lonely.psu = "adjust")
data_final_nest<-
  wi_edu_2010_2020 %>% 
  mutate(
    hv005= hv005 / 1000000,
    
    wealth_tercile = cut(hv271,
                breaks = quantile(hv271, probs = seq(0, 1, 1/3), na.rm = TRUE),
                include.lowest = TRUE,
                labels = c("Bajo","Medio","Alto")),
    
    wealth_tercile = factor(
  ifelse(is.na(wealth_tercile), "Bajo", as.character(wealth_tercile)),
  levels = c("Bajo","Medio","Alto")
)
  ) %>% 
  group_by(year) %>% 
  nest() %>% 
  mutate(
    svydesign_endes= map(.x=data,
                         .f = ~svydesign(ids = ~hv001, strata = ~hv022, 
                                         weights = ~hv005, data = .x, nest = T)),
    
    wi_directa = map(.x = svydesign_endes,
                     .f = ~svyby(~hv270, ~ubigeo, design = .x, svymean, keep.var = T)),
    
    edu_directa = map(.x = svydesign_endes,
                      .f = ~svyby(~hv109, ~ubigeo, design = .x, svymean, keep.var = T)),
    
    
    
    wi_directa_recat = map(.x = svydesign_endes,
                           .f = ~svyby(~hv270_recat, ~ubigeo, design = .x, svymean, keep.var = T)),
    
    
    wi_directa_recat_ter = map(.x = svydesign_endes,
                           .f = ~svyby(~wealth_tercile, ~ubigeo, design = .x, svymean, keep.var = T)),
    
    
    edu_directa_recat = map(.x = svydesign_endes,
                            .f = ~svyby(~hv109_recat, ~ubigeo, design = .x, svymean, keep.var = T))
  )

```

# 4. Merging PCA and full census datasets with direct estimations
Since the datasets may contain values of 0 and 1, and we will transform proportions to the logit form,
we applied a correction: if the direct estimation is equal to 0, we add a small epsilon value. If the proportion is equal to 1, 
we subtract an epsilon value.
```{r}
epsilon <- 0.01


direct_estimation_censo_total <- 
  data_final_nest %>%
  select(year,edu_directa,wi_directa,edu_directa_recat,wi_directa_recat,wi_directa_recat_ter) %>% 
  mutate(
    
    edu_censo_cpa = map(.x = edu_directa,
                    .f = ~distritos %>% 
                      full_join(
                        censo_pca_20 %>% 
                          left_join(.x) %>% 
                          
                          mutate(
                            # MODIFICADO: Corrección epsilon con tolerancia numérica
                            across(.cols = hv109NE:hv109SuE, 
                                   .f = ~case_when(
                                     .x <= epsilon ~ epsilon,
                                     .x >= (1 - epsilon) ~ 1 - epsilon,
                                     TRUE ~ .x
                                   ), 
                                   .names = '{.col}_cor'),
                            
                            # MODIFICADO: Logit manual sin función logit()
                            across(.cols = hv109NE_cor:hv109SuE_cor, 
                                   .f = ~log(.x / (1 - .x)), 
                                   .names = '{.col}_logit'),
                            
                            # Varianza en escala logit
                            var_logit_hv109NE = (se.hv109NE^2)/(hv109NE_cor^2 * (1-hv109NE_cor)^2),
                            var_logit_hv109PE = (se.hv109PE^2)/(hv109PE_cor^2 * (1-hv109PE_cor)^2),
                            var_logit_hv109SE = (se.hv109SE^2)/(hv109SE_cor^2 * (1-hv109SE_cor)^2),
                            var_logit_hv109SuE = (se.hv109SuE^2)/(hv109SuE_cor^2 * (1-hv109SuE_cor)^2),
                            
                            # SE en escala logit (mantener nombres originales)
                            se_hv109NE_cor = var_logit_hv109NE,
                            se_hv109PE_cor = var_logit_hv109PE,
                            se_hv109SE_cor = var_logit_hv109SE,
                            se_hv109SuE_cor = var_logit_hv109SuE
                          ) %>% 
                          
                          mutate(
                            # Floor para varianzas muy pequeñas
                            across(c(var_logit_hv109NE:var_logit_hv109SuE, se_hv109NE_cor:se_hv109SuE_cor), 
                                   .f = ~ifelse(.x < 1e-6, 1e-6, .x)),
                            
                            # Precisión (inversa de varianza)
                            prec_logit_hv109NE = 1 / var_logit_hv109NE,
                            prec_logit_hv109PE = 1 / var_logit_hv109PE,
                            prec_logit_hv109SE = 1 / var_logit_hv109SE,
                            prec_logit_hv109SuE = 1 / var_logit_hv109SuE
                          )
                      
                      ) %>% 
                      group_by(ubigeo) %>% 
                      mutate(
                        id.sp = cur_group_id()
                      ) %>% 
                      ungroup()),
    
    
    
    
    wi_censo_cpa= map(.x = wi_directa,
                   .f = ~distritos %>% 
                     full_join(
                       censo_pca_20 %>% 
                         left_join(.x) %>% 
                         
                         mutate(
                           # MODIFICADO: Corrección epsilon con tolerancia numérica
                           across(.cols = hv2701st:hv2705th, 
                                  .f = ~case_when(
                                    .x <= epsilon ~ epsilon,
                                    .x >= (1 - epsilon) ~ 1 - epsilon,
                                    TRUE ~ .x
                                  ), 
                                  .names = '{.col}_cor'),
                           
                           # MODIFICADO: Logit manual sin función logit()
                           across(.cols = hv2701st_cor:hv2705th_cor, 
                                  .f = ~log(.x / (1 - .x)), 
                                  .names = '{.col}_logit'),
                           
                           # Varianza en escala logit
                           var_logit_hv2701st = (se.hv2701st^2)/(hv2701st_cor^2 * (1-hv2701st_cor)^2),
                           var_logit_hv2702nd = (se.hv2702nd^2)/(hv2702nd_cor^2 * (1-hv2702nd_cor)^2),
                           var_logit_hv2703rd = (se.hv2703rd^2)/(hv2703rd_cor^2 * (1-hv2703rd_cor)^2),
                           var_logit_hv2704th = (se.hv2704th^2)/(hv2704th_cor^2 * (1-hv2704th_cor)^2),
                           var_logit_hv2705th = (se.hv2705th^2)/(hv2705th_cor^2 * (1-hv2705th_cor)^2),
                           
                           # SE en escala logit (mantener nombres originales)
                           se_hv2701st_cor = var_logit_hv2701st,
                           se_hv2702nd_cor = var_logit_hv2702nd,
                           se_hv2703rd_cor = var_logit_hv2703rd,
                           se_hv2704th_cor = var_logit_hv2704th,
                           se_hv2705th_cor = var_logit_hv2705th
                         ) %>% 
                         
                         mutate(
                           # Floor para varianzas muy pequeñas
                           across(c(var_logit_hv2701st:var_logit_hv2705th, se_hv2701st_cor:se_hv2705th_cor), 
                                  .f = ~ifelse(.x < 1e-6, 1e-6, .x)),
                           
                           # Precisión (inversa de varianza)
                           prec_logit_hv2701st = 1 / var_logit_hv2701st,
                           prec_logit_hv2702nd = 1 / var_logit_hv2702nd,
                           prec_logit_hv2703rd = 1 / var_logit_hv2703rd,
                           prec_logit_hv2704th = 1 / var_logit_hv2704th,
                           prec_logit_hv2705th = 1 / var_logit_hv2705th
                         )
                       
                     ) %>% 
                     group_by(ubigeo) %>% 
                     mutate(
                       id.sp = cur_group_id()
                     ) %>% 
                     ungroup()),
    
    
    edu_censo_cpa_recat = map(.x = edu_directa_recat,
                        .f = ~distritos %>% 
                          full_join(
                            censo_pca_20 %>% 
                              left_join(.x) %>% 
                              
                              mutate(
                                # MODIFICADO: Corrección epsilon con tolerancia numérica
                                across(.cols = hv109_recat, 
                                       .f = ~case_when(
                                         .x <= epsilon ~ epsilon,
                                         .x >= (1 - epsilon) ~ 1 - epsilon,
                                         TRUE ~ .x
                                       ), 
                                       .names = '{.col}_cor'),
                                
                                # MODIFICADO: Logit manual sin función logit()
                                across(.cols = hv109_recat_cor, 
                                       .f = ~log(.x / (1 - .x)), 
                                       .names = '{.col}_logit'),
                                
                                # Varianza en escala logit
                                var_logit_hv109_recat = (se^2)/(hv109_recat_cor^2 * (1-hv109_recat_cor)^2),
                                
                                # SE en escala logit (mantener nombre original)
                                se_hv109_recat_cor = se
                              ) %>% 
                              
                              mutate(
                                # Floor para varianzas muy pequeñas
                                across(c(var_logit_hv109_recat, se_hv109_recat_cor), 
                                       .f = ~ifelse(.x < 1e-6, 1e-6, .x)),
                                
                                # Precisión (inversa de varianza)
                                prec_logit_hv109_recat = 1 / var_logit_hv109_recat
                              )
                            
                          ) %>% 
                          group_by(ubigeo) %>% 
                          mutate(
                            id.sp = cur_group_id()
                          ) %>% 
                          ungroup()),
    
    
    wi_censo_cpa_recat= map(.x = wi_directa_recat,
                      .f = ~distritos %>% 
                        full_join(
                          censo_pca_20 %>% 
                            left_join(.x) %>% 
                            
                            mutate(
                              # MODIFICADO: Corrección epsilon con tolerancia numérica
                              across(.cols = hv270_recat, 
                                     .f = ~case_when(
                                       .x <= epsilon ~ epsilon,
                                       .x >= (1 - epsilon) ~ 1 - epsilon,
                                       TRUE ~ .x
                                     ), 
                                     .names = '{.col}_cor'),
                              
                              # MODIFICADO: Logit manual sin función logit()
                              across(.cols = hv270_recat_cor, 
                                     .f = ~log(.x / (1 - .x)), 
                                     .names = '{.col}_logit'),
                              
                              # Varianza en escala logit
                              var_logit_hv270_recat = (se^2)/(hv270_recat_cor^2 * (1-hv270_recat_cor)^2),
                              
                              # SE en escala logit (mantener nombre original)
                              se_hv270_recat_cor = se
                            ) %>% 
                            
                            mutate(
                              # Floor para varianzas muy pequeñas
                              across(c(var_logit_hv270_recat, se_hv270_recat_cor), 
                                     .f = ~ifelse(.x < 1e-6, 1e-6, .x)),
                              
                              # Precisión (inversa de varianza)
                              prec_logit_hv270_recat = 1 / var_logit_hv270_recat
                            )
                          
                        ) %>% 
                        group_by(ubigeo) %>% 
                        mutate(
                          id.sp = cur_group_id()
                        ) %>% 
                        ungroup()),
    
    
    
    wi_censo_cpa_recat_ter= map(.x = wi_directa_recat_ter ,
                      .f = ~distritos %>% 
                        full_join(
                          censo_pca_20 %>% 
                            left_join(.x) %>% 
                            
                            mutate(
                              # MODIFICADO: Corrección epsilon con tolerancia numérica
                              across(.cols = wealth_tercileBajo, 
                                     .f = ~case_when(
                                       .x <= epsilon ~ epsilon,
                                       .x >= (1 - epsilon) ~ 1 - epsilon,
                                       TRUE ~ .x
                                     ), 
                                     .names = '{.col}_cor'),
                              
                              # MODIFICADO: Logit manual sin función logit()
                              across(.cols = wealth_tercileBajo_cor, 
                                     .f = ~log(.x / (1 - .x)), 
                                     .names = '{.col}_logit'),
                              
                              # Varianza en escala logit
                              var_wealth_tercileBajo_logit = (se.wealth_tercileBajo^2)/(wealth_tercileBajo_cor^2 * (1-wealth_tercileBajo_cor)^2),
                              
                              # SE en escala logit (mantener nombre original)
                              se_wealth_tercileBajo_logit = sqrt(var_wealth_tercileBajo_logit)
                            ) 
                          
                        ) %>% 
                        group_by(ubigeo) %>% 
                        mutate(
                          id.sp = cur_group_id()
                        ) %>% 
                        ungroup()),
    
    
    
    
    edu_censo_total = map(.x = edu_directa,
                          .f = ~distritos %>% 
                            full_join(
                              censo_final_loreto %>%
                                left_join(.x) %>% 
                                
                                mutate(
                                  # MODIFICADO: Corrección epsilon con tolerancia numérica
                                  across(.cols = hv109NE:hv109SuE, 
                                         .f = ~case_when(
                                           .x <= epsilon ~ epsilon,
                                           .x >= (1 - epsilon) ~ 1 - epsilon,
                                           TRUE ~ .x
                                         ), 
                                         .names = '{.col}_cor'),
                                  
                                  # MODIFICADO: Logit manual sin función logit()
                                  across(.cols = hv109NE_cor:hv109SuE_cor, 
                                         .f = ~log(.x / (1 - .x)), 
                                         .names = '{.col}_logit'),
                                  
                                  # Varianza en escala logit
                                  var_logit_hv109NE = (se.hv109NE^2)/(hv109NE_cor^2 * (1-hv109NE_cor)^2),
                                  var_logit_hv109PE = (se.hv109PE^2)/(hv109PE_cor^2 * (1-hv109PE_cor)^2),
                                  var_logit_hv109SE = (se.hv109SE^2)/(hv109SE_cor^2 * (1-hv109SE_cor)^2),
                                  var_logit_hv109SuE = (se.hv109SuE^2)/(hv109SuE_cor^2 * (1-hv109SuE_cor)^2),
                                  
                                  # SE en escala logit (mantener nombres originales)
                                  se_hv109NE_cor = var_logit_hv109NE,
                                  se_hv109PE_cor = var_logit_hv109PE,
                                  se_hv109SE_cor = var_logit_hv109SE,
                                  se_hv109SuE_cor = var_logit_hv109SuE
                                ) %>% 
                                
                                mutate(
                                  # Floor para varianzas muy pequeñas
                                  across(c(var_logit_hv109NE:var_logit_hv109SuE, se_hv109NE_cor:se_hv109SuE_cor), 
                                         .f = ~ifelse(.x < 1e-6, 1e-6, .x)),
                                  
                                  # Precisión (inversa de varianza)
                                  prec_logit_hv109NE = 1 / var_logit_hv109NE,
                                  prec_logit_hv109PE = 1 / var_logit_hv109PE,
                                  prec_logit_hv109SE = 1 / var_logit_hv109SE,
                                  prec_logit_hv109SuE = 1 / var_logit_hv109SuE
                                )
                              
                            ) %>% 
                            group_by(ubigeo) %>% 
                            mutate(
                              id.sp = cur_group_id()
                            ) %>% 
                            ungroup()),
    
    
    
    
    wi_censo_total = map(.x = wi_directa,
                         .f = ~distritos %>% 
                           full_join(
                             censo_final_loreto %>% 
                               left_join(.x) %>% 
                               
                               mutate(
                                 # MODIFICADO: Corrección epsilon con tolerancia numérica
                                 across(.cols = hv2701st:hv2705th, 
                                        .f = ~case_when(
                                          .x <= epsilon ~ epsilon,
                                          .x >= (1 - epsilon) ~ 1 - epsilon,
                                          TRUE ~ .x
                                        ), 
                                        .names = '{.col}_cor'),
                                 
                                 # MODIFICADO: Logit manual sin función logit()
                                 across(.cols = hv2701st_cor:hv2705th_cor, 
                                        .f = ~log(.x / (1 - .x)), 
                                        .names = '{.col}_logit'),
                                 
                                 # Varianza en escala logit
                                 var_logit_hv2701st = (se.hv2701st^2)/(hv2701st_cor^2 * (1-hv2701st_cor)^2),
                                 var_logit_hv2702nd = (se.hv2702nd^2)/(hv2702nd_cor^2 * (1-hv2702nd_cor)^2),
                                 var_logit_hv2703rd = (se.hv2703rd^2)/(hv2703rd_cor^2 * (1-hv2703rd_cor)^2),
                                 var_logit_hv2704th = (se.hv2704th^2)/(hv2704th_cor^2 * (1-hv2704th_cor)^2),
                                 var_logit_hv2705th = (se.hv2705th^2)/(hv2705th_cor^2 * (1-hv2705th_cor)^2),
                                 
                                 # SE en escala logit (mantener nombres originales)
                                 se_hv2701st_cor = var_logit_hv2701st,
                                 se_hv2702nd_cor = var_logit_hv2702nd,
                                 se_hv2703rd_cor = var_logit_hv2703rd,
                                 se_hv2704th_cor = var_logit_hv2704th,
                                 se_hv2705th_cor = var_logit_hv2705th
                               ) %>% 
                               
                               mutate(
                                 # Floor para varianzas muy pequeñas
                                 across(c(var_logit_hv2701st:var_logit_hv2705th, se_hv2701st_cor:se_hv2705th_cor), 
                                        .f = ~ifelse(.x < 1e-6, 1e-6, .x)),
                                 
                                 # Precisión (inversa de varianza)
                                 prec_logit_hv2701st = 1 / var_logit_hv2701st,
                                 prec_logit_hv2702nd = 1 / var_logit_hv2702nd,
                                 prec_logit_hv2703rd = 1 / var_logit_hv2703rd,
                                 prec_logit_hv2704th = 1 / var_logit_hv2704th,
                                 prec_logit_hv2705th = 1 / var_logit_hv2705th
                               )
                             
                           ) %>% 
                           group_by(ubigeo) %>% 
                           mutate(
                             id.sp = cur_group_id()
                           ) %>% 
                           ungroup())
  )
```


# 5. save datasets
```{r}
edu_censo_final=direct_estimation_censo_total %>% select(year,edu_censo_cpa) %>% 
  unnest(edu_censo_cpa) %>% ungroup() %>% select(-geometry)

wi_censo_final=direct_estimation_censo_total %>% select(year,wi_censo_cpa) %>% 
  unnest(wi_censo_cpa) %>% ungroup() %>% select(-geometry)

edu_censo_final_recat=direct_estimation_censo_total %>% select(year,edu_censo_cpa_recat) %>%
  unnest(edu_censo_cpa_recat) %>% ungroup() %>% select(-geometry)

wi_censo_final_recat=direct_estimation_censo_total %>% select(year,wi_censo_cpa_recat) %>% 
  unnest(wi_censo_cpa_recat) %>% ungroup() %>% select(-geometry)


wi_censo_final_recat_tercil=direct_estimation_censo_total %>% select(year,wi_censo_cpa_recat_ter) %>% 
  unnest(wi_censo_cpa_recat_ter) %>% ungroup() %>% select(-geometry)

#edu_censo_total=direct_estimation_censo_total %>% select(year,edu_censo_total) %>% unnest(edu_censo_total) %>% ungroup() %>% select(-geometry)
# wi_censo_total=direct_estimation_censo_total %>% select(year,wi_censo_total) %>% unnest(wi_censo_total) %>% ungroup() %>% select(-geometry)
```


```{r, echo=FALSE}
write.csv(edu_censo_final, "./data/edu_censo_final.csv", row.names = F)
write.csv(wi_censo_final, "./data/wi_censo_final.csv", row.names = F)

write.csv(edu_censo_final_recat, "./data/edu_censo_final_recat.csv", row.names = F)
write.csv(wi_censo_final_recat, "./data/wi_censo_final_recat.csv", row.names = F)
write.csv(wi_censo_final_recat_tercil, "./data/wi_censo_final_recat_tercil.csv", row.names = F)

#write.csv(edu_censo_total, "./data/edu_censo_total.csv", row.names = F)
#write.csv(wi_censo_total, "./data/wi_censo_total.csv", row.names = F)

```

```{r}
year = c(2010:2020)
df_final_edu <- map2_df(data_final_nest$edu_directa_recat, year, ~ mutate(.x, year = .y)) %>% 
  as_tibble() %>% 
  rename(
    estimate = hv109_recat
  ) %>% 
  mutate(
    index = "edu"
  ) %>% 
  select(year,ubigeo,index,everything())

df_final_wi <- map2_df(data_final_nest$wi_directa_recat, year, ~ mutate(.x, year = .y)) %>% 
  as_tibble() %>% 
  rename(
    estimate = hv270_recat
  ) %>% 
  mutate(
    index = "wi"
  ) %>% 
  select(year,ubigeo,index,everything())

df_final<-
  bind_rows(df_final_edu,df_final_wi) %>% 
  filter(year %in% c(2010,2012,2014,2016,2018,2020))

write.csv(df_final,"./data/aux_data/fig1_data.csv",row.names = F)

df_final_c<-
  bind_rows(df_final_edu,df_final_wi) 

write.csv(df_final_c,"./data/aux_data/fig1_data_complt.csv",row.names = F)
  
```

