---
title: "Untitled"
output: html_document
date: "2025-05-28"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = F, message = F, fig.width = 12, fig.height = 8)
```


```{r}
library(tidyverse)
library(janitor)
library(sf)
library(innovar)
library(factoextra)
library(spdep)
library(INLA)
library(car)
library(yardstick)
library(gt)
library(viridis)
```


```{r}
data("Peru")
distritos<- Peru %>% filter(dep=="LORETO") %>% select(ubigeo,geometry)
aux_data<- read.csv("./data/aux_data/aux_data.csv") %>% mutate(ubigeo = as.character(ubigeo))
```

# Data de trabajo
```{r}
edu_censo_final<- read.csv("./data/edu_censo_final_recat.csv") %>% as_tibble() %>% mutate(ubigeo = as.character(ubigeo))

wi_censo_final<- read.csv("./data/wi_censo_final_recat.csv") %>% as_tibble() %>% mutate(ubigeo = as.character(ubigeo))
```

to this point: we have the direct estimates and standard errors at the district level for the wealth index (hv270) 
and education level (hv109). Additionally, due to the complex sampling design of the ENDES, data is available for some
of the 53 districts. On the other hand, variables of interest from the 2017 Census were used as covariates. A total of 92 variables were obtained. 
A PCA was performed, and 20 principal components were selected. 
INLA-based models will be fitted to produce SAE and to predict values for the missing districts.

# Fit INLA

# INLA - Education

## adding auxiliary data

```{r}
edu_censo_final<-
  edu_censo_final %>% 
  left_join(aux_data)
```


```{r}
## HV109 count

## Adding total population to the dataset (using malaria incidence data processed in the main repo)

pop_dist <- read.csv("./data/aux_data/malaria_dist_total.csv") %>% 
  select(year,ubigeo,pop_landscan) %>% 
  #rename(ubigeo = iddist) %>% 
  mutate(
    ubigeo = as.character(ubigeo)
  )
```


```{r}
edu_censo_final2<-
  edu_censo_final %>% 
  #select(year,ubigeo,hv270_recat,hv270_recat_cor) %>% 
  inner_join(pop_dist) %>% 
  mutate(
    pop_landscan = round(pop_landscan),
    cases_number = round(hv109_recat_cor*pop_landscan),
    year_re = as.integer(factor(year)), # for interaction spatial - time
    id.sp2 = id.sp
  )
```



# Spatial component

```{r}
loreto.dist <- poly2nb(distritos)
w.loreto <- nb2mat(loreto.dist, style = "W")
```

# fit models
## comparing family: binomial vs poisson vs negative binomial

```{r}
model_binom <- inla(
  formula = cases_number ~ 1 + 
    f(id.sp, model = "bym", graph = w.loreto) +   # efecto espacial
    f(year, model = "ar1"),                       # efecto temporal
  
  data = edu_censo_final2,
  family = "binomial",
  Ntrials = pop_landscan,
  
  control.predictor = list(compute = TRUE, link = 1),
  control.compute = list(dic = TRUE, waic = TRUE, cpo = TRUE)
)

summary(model_binom)


model_poisson <- inla(
  formula = cases_number ~ 1 + 
    f(id.sp, model = "bym", graph = w.loreto) +   # efecto espacial
    f(year, model = "ar1"),                       # efecto temporal
  
  data = edu_censo_final2,
  family = "poisson",
  offset = log(pop_landscan),  
  
  control.predictor = list(compute = TRUE, link = 1),
  control.compute = list(dic = TRUE, waic = TRUE, cpo = TRUE)
)

summary(model_poisson)

model_nb <- inla(
  formula = cases_number ~ 1 + 
    f(id.sp, model = "bym", graph = w.loreto) +   # efecto espacial
    f(year, model = "ar1"),                       # efecto temporal
  
  data = edu_censo_final2,
  family = "nbinomial",
  offset = log(pop_landscan),  
  
  control.predictor = list(compute = TRUE, link = 1),
  control.compute = list(dic = TRUE, waic = TRUE, cpo = TRUE)
)

summary(model_nb) # The negative binomial family was chosen.
```

# fit models (2) ----

```{r}
modelo_fh10_spatial_edu_1 <- inla(
  
  formula =  cases_number ~ 1 + pc1 + pc2 + pc3 + pc4 + pc5+ 
    pc6 + pc7 + pc8 + pc9 + pc10  +
f(id.sp, model = "bym2", graph = w.loreto, group = year_re,
      control.group = list(model = "ar1")),
  
  data = edu_censo_final2,
  family = "nbinomial",
  offset = log(pop_landscan),  
  
  control.predictor = list(compute = TRUE, link = 1),
  control.compute = list(dic = TRUE, cpo = TRUE, waic = TRUE, config = TRUE)
)
```


```{r}
modelo_fh10_spatial_edu_2 <- inla(
  
  formula =  cases_number ~ 1 + pc1 + pc2 + pc3 + pc4 + pc5+ 
    pc6 + pc7 + pc8 + pc9 + pc10  +
    f(id.sp, model = "bym2", graph = w.loreto) +
    f(year, model = "ar1") +
    f(id.sp2, model = "bym2", graph = w.loreto, group = year_re,
      control.group = list(model = "ar1")),
  
  data = edu_censo_final2,
  family = "nbinomial",
  offset = log(pop_landscan),  
  
  control.predictor = list(compute = TRUE, link = 1),
  control.compute = list(dic = TRUE, cpo = TRUE, waic = TRUE, config = TRUE)
)
```


```{r}
modelo_fh8_spatial_edu_1 <- inla(
  
  formula =  cases_number ~ 1 + pc1 + pc2 + pc3 + pc4 + pc5+ 
    pc6 + pc7 + pc8  +
    f(id.sp, model = "bym2", graph = w.loreto, group = year_re,
      control.group = list(model = "ar1")),
  
  data = edu_censo_final2,
  family = "nbinomial",
  offset = log(pop_landscan),  
  
  control.predictor = list(compute = TRUE, link = 1),
  control.compute = list(dic = TRUE, cpo = TRUE, waic = TRUE, config = TRUE)
)
```


```{r}
modelo_fh8_spatial_edu_2 <- inla(
  
  formula =  cases_number ~ 1 + pc1 + pc2 + pc3 + pc4 + pc5+ 
    pc6 + pc7 + pc8  +
    f(id.sp, model = "bym2", graph = w.loreto) +
    f(year, model = "ar1") +
    f(id.sp2, model = "bym2", graph = w.loreto, group = year_re,
      control.group = list(model = "ar1")),
  
  data = edu_censo_final2,
  family = "nbinomial",
  offset = log(pop_landscan),  
  
  control.predictor = list(compute = TRUE, link = 1),
  control.compute = list(dic = TRUE, cpo = TRUE, waic = TRUE, config = TRUE)
)
```


```{r}
modelo_fh5_spatial_edu_1 <- inla(
  
  formula =  cases_number ~ 1 + pc1 + pc2 + pc3 + pc4 + pc5 + 
f(id.sp, model = "bym2", graph = w.loreto, group = year_re,
      control.group = list(model = "ar1")),
  
  data = edu_censo_final2,
  family = "nbinomial",
  offset = log(pop_landscan),  
  
  control.predictor = list(compute = TRUE, link = 1),
  control.compute = list(dic = TRUE, cpo = TRUE, waic = TRUE, config = TRUE)
)
```


```{r}
modelo_fh5_spatial_edu_2 <- inla(
  
  formula =  cases_number ~ 1 + pc1 + pc2 + pc3 + pc4 + pc5 + 
    f(id.sp, model = "bym2", graph = w.loreto) +
    f(year, model = "ar1") +
    f(id.sp2, model = "bym2", graph = w.loreto, group = year_re,
      control.group = list(model = "ar1")),
  
  data = edu_censo_final2,
  family = "nbinomial",
  offset = log(pop_landscan),  
  
  control.predictor = list(compute = TRUE, link = 1),
  control.compute = list(dic = TRUE, cpo = TRUE, waic = TRUE, config = TRUE)
)
```


```{r}
modelo_fh2_spatial_edu_1 <- inla(
  
  formula =  cases_number ~ 1 + pc1 + pc2 +
f(id.sp, model = "bym2", graph = w.loreto, group = year_re,
      control.group = list(model = "ar1")),
  
  data = edu_censo_final2,
  family = "nbinomial",
  offset = log(pop_landscan),  
  
  control.predictor = list(compute = TRUE, link = 1),
  control.compute = list(dic = TRUE, cpo = TRUE, waic = TRUE, config = TRUE)
)
```


```{r}
modelo_fh2_spatial_edu_2 <- inla(
  
  formula =  cases_number ~ 1 + pc1 + pc2 +
    f(id.sp, model = "bym2", graph = w.loreto) +
    f(year, model = "ar1") +
    f(id.sp2, model = "bym2", graph = w.loreto, group = year_re,
      control.group = list(model = "ar1")),
  
  data = edu_censo_final2,
  family = "nbinomial",
  offset = log(pop_landscan),  
  
  control.predictor = list(compute = TRUE, link = 1),
  control.compute = list(dic = TRUE, cpo = TRUE, waic = TRUE, config = TRUE)
)
```


```{r}

modelo_fh2_spatial_edu_3 <- inla(
  
  formula =  cases_number ~ 1 + pc1 + pc2 + total_millones + edu_dist_mat +
    f(id.sp, model = "bym2", graph = w.loreto, group = year_re,
      control.group = list(model = "ar1")),
  
  data = edu_censo_final2,
  family = "nbinomial",
  offset = log(pop_landscan),  
  
  control.predictor = list(compute = TRUE, link = 1),
  control.compute = list(dic = TRUE, cpo = TRUE, waic = TRUE, config = TRUE)
)
```


```{r}
modelo_fh2_spatial_edu_4 <- inla(
  
  formula =  cases_number ~ 1 + pc1 + pc2 + total_millones + edu_dist_mat +
    f(id.sp, model = "bym2", graph = w.loreto) +
    f(year, model = "ar1") +
    f(id.sp2, model = "bym2", graph = w.loreto, group = year_re,
      control.group = list(model = "ar1")),
  
  data = edu_censo_final2,
  family = "nbinomial",
  offset = log(pop_landscan),  
  
  control.predictor = list(compute = TRUE, link = 1),
  control.compute = list(dic = TRUE, cpo = TRUE, waic = TRUE, config = TRUE)
)
```


# Summary

```{r}
model_metrics <- tibble(
  Modelo = c("Modelo 10-1", "Modelo 8-1",
             "Modelo 5-1", "Modelo 2-1",
             "Modelo 2-2", "Modelo 10-2",
             "Modelo 8-2","Modelo 5-2",
             "Modelo 2-3","Modelo 2-4"),
  DIC     = c(modelo_fh10_spatial_edu_1$dic$dic, modelo_fh8_spatial_edu_1$dic$dic, 
              modelo_fh5_spatial_edu_1$dic$dic, modelo_fh2_spatial_edu_1$dic$dic,modelo_fh2_spatial_edu_2$dic$dic,
              modelo_fh10_spatial_edu_2$dic$dic, modelo_fh8_spatial_edu_2$dic$dic, 
              modelo_fh5_spatial_edu_2$dic$dic, modelo_fh2_spatial_edu_3$dic$dic,modelo_fh2_spatial_edu_4$dic$dic),
  
  pD      = c(modelo_fh10_spatial_edu_1$dic$p.eff, modelo_fh8_spatial_edu_1$dic$p.eff, 
              modelo_fh5_spatial_edu_1$dic$p.eff, modelo_fh2_spatial_edu_1$dic$p.eff,modelo_fh2_spatial_edu_2$dic$p.eff,
              modelo_fh10_spatial_edu_2$dic$p.eff, modelo_fh8_spatial_edu_2$dic$p.eff, 
              modelo_fh5_spatial_edu_2$dic$p.eff, modelo_fh2_spatial_edu_3$dic$p.eff,modelo_fh2_spatial_edu_4$dic$p.eff),
  
  WAIC    = c(modelo_fh10_spatial_edu_1$waic$waic, modelo_fh8_spatial_edu_1$waic$waic, 
              modelo_fh5_spatial_edu_1$waic$waic, modelo_fh2_spatial_edu_1$waic$waic,modelo_fh2_spatial_edu_2$waic$waic,
              modelo_fh10_spatial_edu_2$waic$waic, modelo_fh8_spatial_edu_2$waic$waic, 
              modelo_fh5_spatial_edu_2$waic$waic, modelo_fh2_spatial_edu_3$waic$waic,modelo_fh2_spatial_edu_4$waic$waic),
  
  pWAIC   = c(modelo_fh10_spatial_edu_1$waic$p.eff, modelo_fh8_spatial_edu_1$waic$p.eff, 
              modelo_fh5_spatial_edu_1$waic$p.eff, modelo_fh2_spatial_edu_1$waic$p.eff,modelo_fh2_spatial_edu_2$waic$p.eff,
              modelo_fh10_spatial_edu_2$waic$p.eff, modelo_fh8_spatial_edu_2$waic$p.eff, 
              modelo_fh5_spatial_edu_2$waic$p.eff, modelo_fh2_spatial_edu_3$waic$p.eff,modelo_fh2_spatial_edu_4$waic$p.eff),
  
  logML   = c(modelo_fh10_spatial_edu_1$mlik[1,1], modelo_fh8_spatial_edu_1$mlik[1,1], 
              modelo_fh5_spatial_edu_1$mlik[1,1], modelo_fh2_spatial_edu_1$mlik[1,1],modelo_fh2_spatial_edu_2$mlik[1,1],
              modelo_fh10_spatial_edu_2$mlik[1,1], modelo_fh8_spatial_edu_2$mlik[1,1], 
              modelo_fh5_spatial_edu_2$mlik[1,1], modelo_fh2_spatial_edu_3$mlik[1,1],modelo_fh2_spatial_edu_4$mlik[1,1])
)
```


```{r}
model_metrics %>%
  arrange(WAIC, logML) %>%
  gt() %>%
  tab_header(title = md("**Comparación de modelos INLA**")) %>%
  fmt_number(columns = c(DIC, pD, WAIC, pWAIC, logML), decimals = 2) %>%
  
  tab_style(
    style = list(
      cell_fill(color = "#D5F5E3"),
      cell_text(weight = "bold")
    ),
    locations = cells_body(rows = Modelo %in% c("Modelo 8-2", "Modelo 2-2"))
  ) %>%
  
  tab_style(
    style = list(
      cell_fill(color = "#FFF3B0"),
      cell_text(weight = "bold")
    ),
    locations = cells_body(
      rows = Modelo %in% c("Modelo 2-4", 
                           "Modelo 10-2", "Modelo 5-2")
    )
  )

```

# Model accuracy (1: total values)

```{r}
distritos_sf_predict<-
  edu_censo_final2 %>% 
  mutate(
fit10_spat = (modelo_fh10_spatial_edu_1$summary.fitted.values$mean),
fit8_spat = (modelo_fh8_spatial_edu_1$summary.fitted.values$mean),
fit5_spat = (modelo_fh5_spatial_edu_1$summary.fitted.values$mean),
fit10_spat_2 =(modelo_fh10_spatial_edu_2$summary.fitted.values$mean),
fit8_spat_2 = (modelo_fh8_spatial_edu_2$summary.fitted.values$mean),
fit5_spat_2 = (modelo_fh5_spatial_edu_2$summary.fitted.values$mean),
fit2_spat = (modelo_fh2_spatial_edu_1$summary.fitted.values$mean),
fit2_spat_2 = (modelo_fh2_spatial_edu_2$summary.fitted.values$mean),
fit2_spat_3 = (modelo_fh2_spatial_edu_3$summary.fitted.values$mean),
fit2_spat_4 = (modelo_fh2_spatial_edu_4$summary.fitted.values$mean),
estimacion_real = (edu_censo_final2$cases_number)
  ) %>% 
  left_join(distritos) %>% 
  st_as_sf()
```


```{r}
map.direct_est <-
  distritos_sf_predict %>% 
  filter(year == 2020) %>% 
  ggplot()+
  geom_sf(data = distritos)+
  geom_sf(aes(fill=estimacion_real))+
  scale_fill_viridis(direction = -1,option = "rocket")+
  theme(
    legend.position = "bottom"
  )
```


```{r}
map.8spat <-
  distritos_sf_predict %>% 
  filter(year == 2020) %>% 
  ggplot()+
  geom_sf(data = distritos)+
  geom_sf(aes(fill=fit8_spat))+
  scale_fill_viridis(direction = -1,option = "rocket")+
  theme(
    legend.position = "bottom"
  )
map.10spat <-
  distritos_sf_predict %>% 
  filter(year == 2020) %>% 
  ggplot()+
  geom_sf(data = distritos)+
  geom_sf(aes(fill=fit10_spat))+
  scale_fill_viridis(direction = -1,option = "rocket")+
  theme(
    legend.position = "bottom"
  )

map.5spat<-
  distritos_sf_predict %>% 
  filter(year == 2020) %>% 
  ggplot()+
  geom_sf(data = distritos)+
  geom_sf(aes(fill=fit5_spat))+
  scale_fill_viridis(direction = -1,option = "rocket")+
  theme(
    legend.position = "bottom"
  )

map.2spat<-
  distritos_sf_predict %>% 
  filter(year == 2020) %>% 
  ggplot()+
  geom_sf(data = distritos)+
  geom_sf(aes(fill=fit2_spat))+
  scale_fill_viridis(direction = -1,option = "rocket")+
  theme(
    legend.position = "bottom"
  )


map.2spat_new<-
  distritos_sf_predict %>% 
  filter(year == 2020) %>% 
  ggplot()+
  geom_sf(data = distritos)+
  geom_sf(aes(fill=fit2_spat_2))+
  scale_fill_viridis(direction = -1,option = "rocket")+
  theme(
    legend.position = "bottom"
  )
```


```{r}
cowplot::plot_grid(map.direct_est,
                   map.10spat,map.8spat,map.5spat,
                   map.2spat,
                   map.2spat_new, nrow = 2)
```

## Comparison DE vs Pred. SAE ----

```{r}
distritos_sf_predict %>% 
  select(year,cases_number,fit2_spat_2,fit2_spat,fit5_spat,fit8_spat,fit10_spat) %>% 
  st_drop_geometry() %>%
  filter(!is.na(cases_number)) %>%
  rename(directo = cases_number) %>%
  pivot_longer(cols = starts_with("fit"),
               names_to = "modelo",
               values_to = "predicho") %>%
  ggplot(aes(x = directo, y = predicho, color = modelo)) +
  geom_point(alpha = 0.7, size = 2) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "gray40") +
  facet_grid(year~modelo) +
  labs(
    x = "Estimación directa",
    y = "Predicción del modelo",
    title = "Comparación de estimaciones directas vs modelos SAE",
    subtitle = "Línea diagonal = ajuste perfecto"
  ) +
  theme_minimal()
```



## Performance metrics ----

```{r}

fitted_vals  <-  
distritos_sf_predict %>% 
  select(ubigeo,fit10_spat:estimacion_real) %>% 
  pivot_longer(cols = c(fit10_spat:fit2_spat_4), values_to = "fit", names_to = "modelo") %>% 
  mutate(
    actual = estimacion_real
  )

```


```{r}
#### Metrics ----
perform.metrics <- yardstick::metric_set(mae,mase,smape,rmse)
perform.metrics.dist <- yardstick::metric_set(mae,smape,rmse)
```


```{r}
##### total metrics----
tbl.yrd.full <-  fitted_vals %>% 
  group_by(modelo) %>%
  perform.metrics(truth = actual, estimate = fit)

##### metrics by districts----
tbl.yrd.dist_count <-  fitted_vals %>% 
  group_by(modelo,ubigeo) %>%
  perform.metrics.dist(truth = actual, estimate = fit)
```



## Table in-sample accuracy metrics ----
```{r}
tbl.yrd.full %>% 
  pivot_wider(id_cols = modelo,
              names_from = .metric,
              values_from = .estimate) %>%
  arrange(smape) %>% 
  gt() %>%
  tab_header(title = md("in-sample accuracy metrics")) %>% 
  tab_style(
    style = list(
      cell_fill(color = "#FFF3B0"),  # amarillo claro
      cell_text(weight = "bold")
    ),
    locations = cells_body(
      rows = modelo %in% c("fit8_spat_2","fit10_spat_2","fit2_spat_2")
    )
  )
```



### mapping the performance metrics
```{r}
distritos %>% 
  inner_join(tbl.yrd.dist_count) %>% 
  filter(.metric=="smape" & modelo %in% c("fit2_spat","fit2_spat_2","fit2_spat_3","fit2_spat_4",
                                          "fit5_spat","fit5_spat_2","fit8_spat","fit8_spat_2",
                                          "fit10_spat","fit10_spat_2")) %>% 
  ggplot() +
  geom_sf(aes(fill=.estimate),lwd=0.1) +
  scale_fill_distiller(palette="Reds",direction=1,name="smape", limits = c(0,30)) +
  facet_wrap(vars(modelo), nrow = 2) +
  theme_linedraw(base_size = 10) +
  theme(strip.text = element_text(face = "bold",size = 20),
        legend.position = "bottom")
```



# cross validation ----
```{r}
cpo.10cp.sp <--2*sum(log(modelo_fh10_spatial_edu_1$cpo$cpo),na.rm = T)
cpo.10cp.sp2 <--2*sum(log(modelo_fh10_spatial_edu_2$cpo$cpo),na.rm = T)
cpo.5cp.sp <--2*sum(log(modelo_fh5_spatial_edu_1$cpo$cpo),na.rm = T)
cpo.5cp.sp2 <--2*sum(log(modelo_fh5_spatial_edu_2$cpo$cpo),na.rm = T)
cpo.8cp.sp <--2*sum(log(modelo_fh8_spatial_edu_1$cpo$cpo),na.rm = T)
cpo.8cp.sp2 <--2*sum(log(modelo_fh8_spatial_edu_2$cpo$cpo),na.rm = T)
cpo.2cp.sp <--2*sum(log(modelo_fh2_spatial_edu_1$cpo$cpo),na.rm = T)
cpo.2cp.sp2 <- -2*sum(log(modelo_fh2_spatial_edu_2$cpo$cpo),na.rm = T)
cpo.2cp.sp3 <- -2*sum(log(modelo_fh2_spatial_edu_3$cpo$cpo),na.rm = T)
cpo.2cp.sp4 <--2*sum(log(modelo_fh2_spatial_edu_4$cpo$cpo),na.rm = T)

data.cpo  <-  list(
                   "modelo10spat" = cpo.10cp.sp,
                   "modelo10spat2" = cpo.10cp.sp2,
                   "modelo8spat" = cpo.8cp.sp,
                   "modelo8spat2" = cpo.8cp.sp2,
                   "modelo5spat" = cpo.5cp.sp,
                   "modelo5spat2" = cpo.5cp.sp2,
                   "modelo2spat" = cpo.2cp.sp,
                   "modelo2spat_2" = cpo.2cp.sp2,
                   "modelo2spat_3" = cpo.2cp.sp3,
                   "modelo2spat_4" = cpo.2cp.sp4) %>% 
  as.data.frame()
```


```{r}
data.cpo %>% 
  pivot_longer(cols=colnames(data.cpo),
               names_to = "model",
               values_to = "CPO")  %>% 
  arrange(CPO) %>% 
  gt() %>%
  tab_header(title = md("LOO-CV")) %>% 
  tab_style(
    style = list(
      cell_fill(color = "#FFF3B0"),  # amarillo claro
      cell_text(weight = "bold")
    ),
    locations = cells_body(
      rows = model %in% c("modelo2spat_2")
    )
  )
```


```{r}
bind_cols(
model_metrics %>% 
  arrange(Modelo),

tbl.yrd.full %>% 
  pivot_wider(id_cols = modelo,
              names_from = .metric,
              values_from = .estimate) %>% 
  arrange(modelo),

data.cpo %>% 
  pivot_longer(cols=colnames(data.cpo),
               names_to = "model",
               values_to = "CPO") %>% 
  arrange(model)
) %>% 
  select(Modelo,DIC,pD,WAIC,pWAIC,logML,mae,mase,smape,rmse,CPO) %>% 
  arrange(smape) %>% 
  gt() %>% 
  tab_header(title = md("**General metrics: model diagnosis, errors, LOO-CV**"),
             subtitle = md("**Count values**" )) %>% 
  tab_style(
    style = list(
      cell_fill(color = "#FFF3B0"),  # amarillo claro
      cell_text(weight = "bold")
    ),
    locations = cells_body(
      rows = Modelo %in% c("Modelo 8-2","Modelo 2-2")
    )
  )
```

# Model accuracy (2: proportions values)

```{r}
distritos_sf_predict_prop<-
  edu_censo_final2 %>% 
  mutate(
    fit10_spat = (modelo_fh10_spatial_edu_1$summary.fitted.values$mean),
    fit8_spat = (modelo_fh8_spatial_edu_1$summary.fitted.values$mean),
    fit5_spat = (modelo_fh5_spatial_edu_1$summary.fitted.values$mean),
    fit10_spat_2 =(modelo_fh10_spatial_edu_2$summary.fitted.values$mean),
    fit8_spat_2 = (modelo_fh8_spatial_edu_2$summary.fitted.values$mean),
    fit5_spat_2 = (modelo_fh5_spatial_edu_2$summary.fitted.values$mean),
    fit2_spat = (modelo_fh2_spatial_edu_1$summary.fitted.values$mean),
    fit2_spat_2 = (modelo_fh2_spatial_edu_2$summary.fitted.values$mean),
    fit2_spat_3 = (modelo_fh2_spatial_edu_3$summary.fitted.values$mean),
    fit2_spat_4 = (modelo_fh2_spatial_edu_4$summary.fitted.values$mean),
    estimacion_real = (edu_censo_final2$cases_number)
  ) %>% 
  left_join(distritos) %>% 
  
  mutate(
    across(.cols = c(fit10_spat:estimacion_real), .f=~round(.x/pop_landscan,2))
  ) %>% 
  st_as_sf()
```

## Comparison DE vs Pred. SAE ----

```{r}
distritos_sf_predict_prop %>% 
  select(year,estimacion_real,fit2_spat_2,fit8_spat_2,fit10_spat_2,fit5_spat_2,fit2_spat) %>% 
  st_drop_geometry() %>%
  filter(!is.na(estimacion_real)) %>%
  rename(directo = estimacion_real) %>%
  pivot_longer(cols = starts_with("fit"),
               names_to = "modelo",
               values_to = "predicho") %>%
  ggplot(aes(x = directo, y = predicho, color = modelo)) +
  geom_point(alpha = 0.7, size = 2) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "gray40") +
  facet_grid(year~modelo) +
  labs(
    x = "Estimación directa",
    y = "Predicción del modelo",
    title = "Comparación de estimaciones directas vs modelos SAE",
    subtitle = "Línea diagonal = ajuste perfecto"
  ) +
  theme_minimal() +
  theme(legend.position = "bottom")
```

## Performance metrics ----

```{r}
fitted_vals  <-  
distritos_sf_predict_prop %>% 
  select(ubigeo,fit10_spat:estimacion_real) %>% 
  pivot_longer(cols = c(fit10_spat:fit2_spat_4), values_to = "fit", names_to = "modelo")

```


```{r}
#### Metrics ----
perform.metrics <- yardstick::metric_set(mae,mase,smape,rmse)
perform.metrics.dist <- yardstick::metric_set(mae,smape,rmse)
```


```{r}
##### total metrics----
tbl.yrd.full <-  fitted_vals %>% 
  group_by(modelo) %>%
  perform.metrics(truth = estimacion_real, estimate = fit)

##### metrics by districts----
tbl.yrd.dist_prop <-  fitted_vals %>% 
  group_by(modelo,ubigeo) %>%
  perform.metrics.dist(truth = estimacion_real, estimate = fit)
```



## Table in-sample accuracy metrics ----

```{r}
tbl.yrd.full %>% 
  pivot_wider(id_cols = modelo,
              names_from = .metric,
              values_from = .estimate) %>%         
  gt() %>%
  tab_header(title = md("in-sample accuracy metrics")) %>% 
  tab_style(
    style = list(
      cell_fill(color = "#FFF3B0"),  # amarillo claro
      cell_text(weight = "bold")
    ),
    locations = cells_body(
      rows = modelo %in% c("fit8_spat_2","fit2_spat_2")
    )
  )
```
### mapping the performance metrics
```{r}
distritos %>% 
  inner_join(tbl.yrd.dist_prop) %>% 
  filter(.metric=="smape" & modelo %in% c("fit8_spat_2","fit2_spat_2","fit10_spat_2",
                                          "fit5_spat_2","fit2_spat")) %>% 
  ggplot() +
  geom_sf(aes(fill=.estimate),lwd=0.1) +
  scale_fill_distiller(palette="Reds",direction=1,name="smape", limits = c(0,30)) +
  facet_wrap(vars(modelo), nrow = 1) +
  theme_minimal(base_size = 8) +
  theme(strip.text = element_text(face = "bold",size = 20),
        legend.position = "bottom")
```

# cross validation ----
```{r}

cpo.10cp.sp <--2*sum(log(modelo_fh10_spatial_edu_1$cpo$cpo),na.rm = T)
cpo.10cp.sp2 <--2*sum(log(modelo_fh10_spatial_edu_2$cpo$cpo),na.rm = T)
cpo.5cp.sp <--2*sum(log(modelo_fh5_spatial_edu_1$cpo$cpo),na.rm = T)
cpo.5cp.sp2 <--2*sum(log(modelo_fh5_spatial_edu_2$cpo$cpo),na.rm = T)
cpo.8cp.sp <--2*sum(log(modelo_fh8_spatial_edu_1$cpo$cpo),na.rm = T)
cpo.8cp.sp2 <--2*sum(log(modelo_fh8_spatial_edu_2$cpo$cpo),na.rm = T)
cpo.2cp.sp <--2*sum(log(modelo_fh2_spatial_edu_1$cpo$cpo),na.rm = T)
cpo.2cp.sp2 <- -2*sum(log(modelo_fh2_spatial_edu_2$cpo$cpo),na.rm = T)
cpo.2cp.sp3 <- -2*sum(log(modelo_fh2_spatial_edu_3$cpo$cpo),na.rm = T)
cpo.2cp.sp4 <--2*sum(log(modelo_fh2_spatial_edu_4$cpo$cpo),na.rm = T)

data.cpo  <-  list(
                   "modelo10spat" = cpo.10cp.sp,
                   "modelo10spat2" = cpo.10cp.sp2,
                   "modelo8spat" = cpo.8cp.sp,
                   "modelo8spat2" = cpo.8cp.sp2,
                   "modelo5spat" = cpo.5cp.sp,
                   "modelo5spat2" = cpo.5cp.sp2,
                   "modelo2spat" = cpo.2cp.sp,
                   "modelo2spat_2" = cpo.2cp.sp2,
                   "modelo2spat_3" = cpo.2cp.sp3,
                   "modelo2spat_4" = cpo.2cp.sp4) %>% 
  as.data.frame()
```


```{r}
data.cpo %>% 
  pivot_longer(cols=colnames(data.cpo),
               names_to = "model",
               values_to = "CPO")  %>% 
  arrange(CPO) %>% 
  gt() %>%
  tab_header(title = md("LOO-CV")) %>% 
  tab_style(
    style = list(
      cell_fill(color = "#FFF3B0"),  # amarillo claro
      cell_text(weight = "bold")
    ),
    locations = cells_body(
      rows = model %in% c("modelo2spat_2")
    )
  )
```


```{r}
bind_cols(
model_metrics %>% 
  arrange(Modelo),

tbl.yrd.full %>% 
  pivot_wider(id_cols = modelo,
              names_from = .metric,
              values_from = .estimate) %>% 
  arrange(modelo),

data.cpo %>% 
  pivot_longer(cols=colnames(data.cpo),
               names_to = "model",
               values_to = "CPO") %>% 
  arrange(model)
) %>% 
  select(Modelo,DIC,pD,WAIC,pWAIC,logML,mae,mase,smape,rmse,CPO) %>% 
  arrange(smape) %>% 
  gt() %>% 
  tab_header(title = md("**General metrics: model diagnosis, errors, LOO-CV**"),
             subtitle = md("**Count values**" )) %>% 
  tab_style(
    style = list(
      cell_fill(color = "#FFF3B0"),  # amarillo claro
      cell_text(weight = "bold")
    ),
    locations = cells_body(
      rows = Modelo %in% c("Modelo 8-2","Modelo 2-2")
    )
  )
```

## Visualization 

#### Real values of edu level

##### Absolute values

```{r}
graficos<-
  distritos_sf_predict %>% 
  st_drop_geometry() %>% 
  select(year,ubigeo,estimacion_real,fit10_spat:fit2_spat_2) %>% 
  pivot_longer(cols = fit10_spat:fit2_spat_2) %>% 
  
  group_by(name) %>% 
  nest() %>% 
  
  mutate(
    data_graph = map(.x = data,
                     .f = ~.x %>% pivot_longer(cols = c(estimacion_real,value))),
    
    graph = map2(.x = data_graph, .y = name,
                .f = ~.x %>% ggplot(aes(x = value, y=ubigeo, col = name))+
                  geom_point(alpha = 0.5)+
                  geom_line(aes(group = ubigeo))+
                  facet_wrap(~year, scales = "free")+
                  ggtitle(name))
  )
```

```{r, warning=F}
graficos$graph[[4]]
```

##### Relative values (proportions)

```{r}
library(ggh4x)

graficos_prop<-
  distritos_sf_predict_prop %>% 
  st_drop_geometry() %>% 
  select(year,ubigeo,pop_landscan,estimacion_real,fit10_spat:fit2_spat_2) %>% 
  pivot_longer(cols = fit10_spat:fit2_spat_2) %>% 
  mutate(
    dif = estimacion_real - value
  ) %>% 
  
  group_by(name) %>% 
  nest() %>% 
  
  mutate(
    data_graph = map(.x = data,
                     .f = ~.x %>% pivot_longer(cols = c(estimacion_real,value))),
    
    graph = map2(.x = data_graph, .y = name,
                .f = ~.x %>% ggplot(aes(x = value, y=ubigeo, col = name))+
                  geom_point(alpha = 0.5, aes(size = pop_landscan))+
                  geom_line(aes(group = ubigeo))+
                  facet_wrap(~year, scales = "free",nrow = 2)+
                  
                  facetted_pos_scales(
                    y = list(
                      `2010` = scale_y_discrete(),
                      `2011` = scale_y_discrete(labels = NULL, breaks = NULL),
                      `2012` = scale_y_discrete(labels = NULL, breaks = NULL),
                      `2013` = scale_y_discrete(labels = NULL, breaks = NULL),
                      `2014` = scale_y_discrete(labels = NULL, breaks = NULL),
                      `2015` = scale_y_discrete(labels = NULL, breaks = NULL),
                      `2016` = scale_y_discrete(),
                      `2017` = scale_y_discrete(labels = NULL, breaks = NULL),
                      `2018` = scale_y_discrete(labels = NULL, breaks = NULL),
                      `2019` = scale_y_discrete(labels = NULL, breaks = NULL),
                      `2020` = scale_y_discrete(labels = NULL, breaks = NULL))
                    ) +
                  theme_bw() +
                  theme(legend.position = "bottom",
                        axis.text.y = element_text(size = 7),
                        panel.grid = element_blank())+
                  ggtitle(name))
  )
```

```{r}
graficos_prop$graph[[5]]
```

#### Calculation proportions with cases predicted

```{r}
prop_sae_predict_edu<-
  distritos_sf_predict %>% 
  select(year,ubigeo,pop_landscan,estimacion_real,fit10_spat,
         fit8_spat,fit5_spat,fit2_spat,fit2_spat_2,geometry) %>% 
  mutate(
    across(.cols = c(estimacion_real:fit2_spat_2), .f=~.x/pop_landscan)
  )
```


```{r}
prop_sae_predict_edu %>% 
  pivot_longer(cols = estimacion_real:fit2_spat_2) %>% 
  mutate(
    name = factor(name, levels = c("estimacion_real", "fit10_spat", "fit8_spat", "fit5_spat", 
                                   "fit2_spat", "fit2_spat_2"))
  ) %>% 
  
  ggplot()+
  geom_sf(aes(fill = value)) +
  scale_fill_viridis(direction = -1,option = "rocket") +
  facet_grid(name~year)
```



### Point and interval estimates of proportions (INLA)

```{r}
modelos <- list( # lista para trabajar en bloque
  modelo_fh8_spatial_2 = modelo_fh8_spatial_edu_2,
  modelo_fh10_spatial_2 = modelo_fh10_spatial_edu_2,
  modelo_fh5_spatial_2 = modelo_fh5_spatial_edu_2,
  modelo_fh2_spatial = modelo_fh2_spatial_edu_1,
  modelo_fh2_spatial_2 = modelo_fh2_spatial_edu_2
)
```


```{r}
resultados_SAE_loreto_edu <- 
  imap_dfr(modelos, ~{ # formato long de las estimaciones y los IC de cada modelo distrito-año
  data.frame(
    year = edu_censo_final$year,
    ubigeo = edu_censo_final$ubigeo,
    mean = .x$summary.fitted.values$mean,
    lower = .x$summary.fitted.values$`0.025quant`,
    upper = .x$summary.fitted.values$`0.975quant`,
    modelo = .y
  )
}) %>% 
  inner_join(pop_dist) %>% 
  rename(
    edu_estimate = mean
  ) %>% 
  mutate(
    across(.cols = edu_estimate:upper, .f=~(.x/pop_landscan), .names = '{.col}_prop')
  )

write.csv(resultados_SAE_loreto_edu, "./data/final_data/resultados_SAE_loreto_edu.csv", row.names = F )
```


```{r}
resultados_SAE_loreto_edu %>% 
  filter(modelo %in% c("modelo_fh8_spatial_2","modelo_fh2_spatial_2")) %>% 
  ggplot(aes(x = edu_estimate_prop, y = ubigeo, col = modelo))+
  geom_pointrange(aes(xmin = lower_prop, xmax = upper_prop))+
  facet_grid(modelo~year)
```


# accuracy metris vs population

## count values
```{r}
tbl.yrd.dist_count %>% 
  full_join(
pop_dist %>% 
  group_by(ubigeo) %>% 
  summarise(pop_model = sum(pop_landscan))
) %>% 
  filter(
    modelo %in% c("estimacion_real", "fit10_spat", "fit8_spat", "fit5_spat", 
                                   "fit2_spat", "fit2_spat_2")
  ) %>% 
  
  ggplot(aes(x = .estimate, y = log(pop_model), col = modelo))+
  geom_point(aes(shape = modelo), alpha = .5, size = 3)+
  geom_line(aes(group = ubigeo)) +
  
  coord_flip() +
  
  facet_wrap(~.metric, scales = "free", ncol = 1)
```

## prop values
```{r}
tbl.yrd.dist_prop %>% 
  full_join(
pop_dist %>% 
  group_by(ubigeo) %>% 
  summarise(pop_model = sum(pop_landscan))
) %>% 
  filter(
    modelo %in% c("estimacion_real", "fit10_spat", "fit8_spat", "fit5_spat", 
                                   "fit2_spat", "fit2_spat_2")
  ) %>% 
  
  ggplot(aes(x = .estimate, y = log(pop_model), col = modelo))+
  geom_point(aes(shape = modelo), alpha = .5, size = 3)+
  geom_line(aes(group = ubigeo)) +
  
  coord_flip() +
  
  facet_wrap(~.metric, scales = "free", ncol = 1)
```

