---
title: "Untitled"
output: html_document
date: "2025-05-28"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = F, message = F, fig.width = 12, fig.height = 8)
```


```{r}
library(tidyverse)
library(janitor)
library(sf)
library(innovar)
library(factoextra)
library(spdep)
library(INLA)
library(car)
library(yardstick)
library(gt)
library(viridis)
```


```{r}
data("Peru")
distritos<- Peru %>% filter(dep=="LORETO") %>% select(ubigeo,geometry)
aux_data<- read.csv("./data/aux_data/aux_data.csv") %>% mutate(ubigeo = as.character(ubigeo))
```

# Data de trabajo
```{r}
edu_censo_final<- read.csv("./data/edu_censo_final_recat.csv") %>% as_tibble() %>% mutate(ubigeo = as.character(ubigeo))

wi_censo_final<- read.csv("./data/wi_censo_final_recat.csv") %>% as_tibble() %>% mutate(ubigeo = as.character(ubigeo))
```

to this point: we have the direct estimates and standard errors at the district level for the wealth index (hv270) 
and education level (hv109). Additionally, due to the complex sampling design of the ENDES, data is available for some
of the 53 districts. On the other hand, variables of interest from the 2017 Census were used as covariates. A total of 92 variables were obtained. 
A PCA was performed, and 20 principal components were selected. 
INLA-based models will be fitted to produce SAE and to predict values for the missing districts.

# Fit INLA

# INLA - Education

## adding auxiliary data

```{r}
edu_censo_final<-
  edu_censo_final %>% 
  left_join(aux_data)
```


```{r}
## HV109 count

## Adding total population to the dataset (using malaria incidence data processed in the main repo)

pop_dist <- read.csv("./data/aux_data/malaria_dist_total.csv") %>% 
  select(year,iddist,pop_landscan) %>% 
  rename(ubigeo = iddist) %>% 
  mutate(
    ubigeo = as.character(ubigeo)
  )
```


```{r}
edu_censo_final2<-
  edu_censo_final %>% 
  #select(year,ubigeo,hv270_recat,hv270_recat_cor) %>% 
  inner_join(pop_dist) %>% 
  mutate(
    pop_landscan = round(pop_landscan),
    cases_number = round(hv109_recat_cor*pop_landscan)
  )
```



# Spatial component

```{r}
loreto.dist <- poly2nb(distritos)
w.loreto <- nb2mat(loreto.dist, style = "W")
```

# fit models
## comparing family: binomial vs poisson vs negative binomial

```{r}
model_binom <- inla(
  formula = cases_number ~ 1 + 
    f(id.sp, model = "bym", graph = w.loreto) +   # efecto espacial
    f(year, model = "ar1"),                       # efecto temporal
  
  data = edu_censo_final2,
  family = "binomial",
  Ntrials = pop_landscan,
  
  control.predictor = list(compute = TRUE, link = 1),
  control.compute = list(dic = TRUE, waic = TRUE, cpo = TRUE)
)

summary(model_binom)


model_poisson <- inla(
  formula = cases_number ~ 1 + 
    f(id.sp, model = "bym", graph = w.loreto) +   # efecto espacial
    f(year, model = "ar1"),                       # efecto temporal
  
  data = edu_censo_final2,
  family = "poisson",
  offset = log(pop_landscan),  
  
  control.predictor = list(compute = TRUE, link = 1),
  control.compute = list(dic = TRUE, waic = TRUE, cpo = TRUE)
)

summary(model_poisson)

model_nb <- inla(
  formula = cases_number ~ 1 + 
    f(id.sp, model = "bym", graph = w.loreto) +   # efecto espacial
    f(year, model = "ar1"),                       # efecto temporal
  
  data = edu_censo_final2,
  family = "nbinomial",
  offset = log(pop_landscan),  
  
  control.predictor = list(compute = TRUE, link = 1),
  control.compute = list(dic = TRUE, waic = TRUE, cpo = TRUE)
)

summary(model_nb) # The negative binomial family was chosen.
```

# fit models (2) ----

```{r}
modelo_fh8_spatial <- inla(
  
  formula =  cases_number ~ 1 + pc1 + pc2 + pc3 + pc4 + pc5+ 
    pc6 + pc7 + pc8  +
    f(id.sp, model = "bym2", graph = w.loreto)+
    f(year, model = "ar1"),
  
  data = edu_censo_final2,
  family = "nbinomial",
  offset = log(pop_landscan),  
  
  control.predictor = list(compute = TRUE, link = 1),
  control.compute = list(dic = TRUE, cpo = TRUE, waic = TRUE)
)
```


```{r}
modelo_fh10_spatial <- inla(
  
  formula =  cases_number ~ 1 + pc1 + pc2 + pc3 + pc4 + pc5+ 
    pc6 + pc7 + pc8 + pc9 + pc10  +
    f(id.sp, model = "bym2", graph = w.loreto)+
    f(year, model = "ar1"),
  
  data = edu_censo_final2,
  family = "nbinomial",
  offset = log(pop_landscan),  
  
  control.predictor = list(compute = TRUE, link = 1),
  control.compute = list(dic = TRUE, cpo = TRUE, waic = TRUE)
)
```


```{r}
modelo_fh5_spatial <- inla(
  
  formula =  cases_number ~ 1 + pc1 + pc2 + pc3 + pc4 + pc5 + 
    f(id.sp, model = "bym2", graph = w.loreto)+ 
    f(year, model = "ar1"),
  
  data = edu_censo_final2,
  family = "nbinomial",
  offset = log(pop_landscan),  
  
  control.predictor = list(compute = TRUE, link = 1),
  control.compute = list(dic = TRUE, cpo = TRUE, waic = TRUE)
)
```


```{r}

modelo_fh2_spatial <- inla(
  
  formula =  cases_number ~ 1 + pc1 + pc2 +
    f(year, model = "ar1") +
    f(id.sp, model = "bym2", graph = w.loreto),
  
  data = edu_censo_final2,
  family = "nbinomial",
  offset = log(pop_landscan),  
  
  control.predictor = list(compute = TRUE, link = 1),
  control.compute = list(dic = TRUE, cpo = TRUE, waic = TRUE)
)
```


```{r}
modelo_fh2_spatial_new <- inla(
  
  formula =  cases_number ~ 1 + pc1 + pc2 + total_millones +
    f(year, model = "ar1") +
    f(id.sp, model = "bym2", graph = w.loreto),
  
  data = edu_censo_final2,
  family = "nbinomial",
  offset = log(pop_landscan),  
  
  control.predictor = list(compute = TRUE, link = 1),
  control.compute = list(dic = TRUE, cpo = TRUE, waic = TRUE)
)
```

# Summary

```{r}
model_metrics <- tibble(
  Modelo = c("Modelo 10", "Modelo 8", "Modelo 5", "Modelo 2","modelo 2 new"),
  DIC     = c(modelo_fh10_spatial$dic$dic, modelo_fh8_spatial$dic$dic, 
              modelo_fh5_spatial$dic$dic, modelo_fh2_spatial$dic$dic,modelo_fh2_spatial_new$dic$dic),
  pD      = c(modelo_fh10_spatial$dic$p.eff, modelo_fh8_spatial$dic$p.eff, 
              modelo_fh5_spatial$dic$p.eff, modelo_fh2_spatial$dic$p.eff,modelo_fh2_spatial_new$dic$p.eff),
  WAIC    = c(modelo_fh10_spatial$waic$waic, modelo_fh8_spatial$waic$waic, 
              modelo_fh5_spatial$waic$waic, modelo_fh2_spatial$waic$waic,modelo_fh2_spatial_new$waic$waic),
  pWAIC   = c(modelo_fh10_spatial$waic$p.eff, modelo_fh8_spatial$waic$p.eff, 
              modelo_fh5_spatial$waic$p.eff, modelo_fh2_spatial$waic$p.eff,modelo_fh2_spatial_new$waic$p.eff),
  logML   = c(modelo_fh10_spatial$mlik[1,1], modelo_fh8_spatial$mlik[1,1], 
              modelo_fh5_spatial$mlik[1,1], modelo_fh2_spatial$mlik[1,1],modelo_fh2_spatial_new$mlik[1,1])
)
```


```{r}
model_metrics %>%
  gt() %>%
  tab_header(title = md("**Comparación de modelos INLA**")) %>%
  fmt_number(columns = c(DIC, pD, WAIC, pWAIC, logML), decimals = 2) %>% 
  
  tab_style(
    style = list(
      cell_fill(color = "#FFF3B0"),  # amarillo claro
      cell_text(weight = "bold")
    ),
    locations = cells_body(
      rows = Modelo %in% c("Modelo 2")
    )
  )
```

# Model accuracy

```{r}
distritos_sf_predict<-
  edu_censo_final2 %>% 
  mutate(
    fit8_spat = (modelo_fh8_spatial$summary.fitted.values$mean),
    fit10_spat = (modelo_fh10_spatial$summary.fitted.values$mean),
    fit5_spat = (modelo_fh5_spatial$summary.fitted.values$mean),
    fit2_spat = (modelo_fh2_spatial$summary.fitted.values$mean),
    fit2_new = (modelo_fh2_spatial_new$summary.fitted.values$mean),
    real_cases = (edu_censo_final2$cases_number)
  ) %>% 
  left_join(distritos) %>% 
  st_as_sf()
```


```{r}
map.direct_est <-
  distritos_sf_predict %>% 
  filter(year == 2020) %>% 
  ggplot()+
  geom_sf(data = distritos)+
  geom_sf(aes(fill=real_cases))+
  scale_fill_viridis(direction = -1,option = "rocket")+
  theme(
    legend.position = "bottom"
  )
```


```{r}
map.8spat <-
  distritos_sf_predict %>% 
  filter(year == 2020) %>% 
  ggplot()+
  geom_sf(data = distritos)+
  geom_sf(aes(fill=fit8_spat))+
  scale_fill_viridis(direction = -1,option = "rocket")+
  theme(
    legend.position = "bottom"
  )
map.10spat <-
  distritos_sf_predict %>% 
  filter(year == 2020) %>% 
  ggplot()+
  geom_sf(data = distritos)+
  geom_sf(aes(fill=fit10_spat))+
  scale_fill_viridis(direction = -1,option = "rocket")+
  theme(
    legend.position = "bottom"
  )

map.5spat<-
  distritos_sf_predict %>% 
  filter(year == 2020) %>% 
  ggplot()+
  geom_sf(data = distritos)+
  geom_sf(aes(fill=fit5_spat))+
  scale_fill_viridis(direction = -1,option = "rocket")+
  theme(
    legend.position = "bottom"
  )

map.2spat<-
  distritos_sf_predict %>% 
  filter(year == 2020) %>% 
  ggplot()+
  geom_sf(data = distritos)+
  geom_sf(aes(fill=fit2_spat))+
  scale_fill_viridis(direction = -1,option = "rocket")+
  theme(
    legend.position = "bottom"
  )


map.2spat_new<-
  distritos_sf_predict %>% 
  filter(year == 2020) %>% 
  ggplot()+
  geom_sf(data = distritos)+
  geom_sf(aes(fill=fit2_new))+
  scale_fill_viridis(direction = -1,option = "rocket")+
  theme(
    legend.position = "bottom"
  )
```


```{r}
cowplot::plot_grid(map.10spat,map.8spat,map.5spat,
                   map.2spat,
                   map.2spat_new, nrow = 2)
```

## Comparison DE vs Pred. SAE ----

```{r}
distritos_sf_predict %>% 
  select(year,cases_number,fit2_new,fit2_spat,fit5_spat,fit8_spat,fit10_spat) %>% 
  st_drop_geometry() %>%
  filter(!is.na(cases_number)) %>%
  rename(directo = cases_number) %>%
  pivot_longer(cols = starts_with("fit"),
               names_to = "modelo",
               values_to = "predicho") %>%
  ggplot(aes(x = directo, y = predicho, color = modelo)) +
  geom_point(alpha = 0.7, size = 2) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "gray40") +
  facet_grid(year~modelo) +
  labs(
    x = "Estimación directa",
    y = "Predicción del modelo",
    title = "Comparación de estimaciones directas vs modelos SAE",
    subtitle = "Línea diagonal = ajuste perfecto"
  ) +
  theme_minimal()
```



## Performance metrics ----

```{r}
fit8_spat <- (modelo_fh8_spatial$summary.fitted.values$mean)
fit10_spat <- (modelo_fh10_spatial$summary.fitted.values$mean)
fit5_spat <- (modelo_fh5_spatial$summary.fitted.values$mean)
fit2_spat <- (modelo_fh2_spatial$summary.fitted.values$mean)
fit2_new <- (modelo_fh2_spatial_new$summary.fitted.values$mean)
estimacion_real <- (edu_censo_final2$cases_number)
```


```{r}
fitted_vals  <-  list(
                      "pc10sp" = fit10_spat,
                      "pc8sp" = fit8_spat,
                      "pc5sp" = fit5_spat,
                      "pc2sp" = fit2_spat,
                      "pc2new" = fit2_new) %>% 
  as.data.frame() %>% 
  gather(key = "modelo", value = "fit") %>% 
  group_by(modelo) %>% 
  mutate(
    actual = estimacion_real,
    ubigeo = as.character(edu_censo_final$ubigeo)
  )
```


```{r}
#### Metrics ----
perform.metrics <- yardstick::metric_set(mae,mase,smape,rmse)
perform.metrics.dist <- yardstick::metric_set(mae,smape,rmse)
```


```{r}
##### total metrics----
tbl.yrd.full <-  fitted_vals %>% 
  group_by(modelo) %>%
  perform.metrics(truth = actual, estimate = fit)

##### metrics by districts----
tbl.yrd.dist <-  fitted_vals %>% 
  group_by(modelo,ubigeo) %>%
  perform.metrics.dist(truth = actual, estimate = fit)
```



## Table in-sample accuracy metrics ----
```{r}
tbl.yrd.full %>% 
  pivot_wider(id_cols = modelo,
              names_from = .metric,
              values_from = .estimate) %>%         
  gt() %>%
  tab_header(title = md("in-sample accuracy metrics")) %>% 
  tab_style(
    style = list(
      cell_fill(color = "#FFF3B0"),  # amarillo claro
      cell_text(weight = "bold")
    ),
    locations = cells_body(
      rows = modelo %in% c("pc2sp")
    )
  )
```



### mapping the performance metrics
```{r}
distritos %>% 
  inner_join(tbl.yrd.dist) %>% 
  filter(.metric=="smape" & modelo %in% c("pc5sp","pc10sp","pc8sp",
                                          "pc2sp","pc2new")) %>% 
  ggplot() +
  geom_sf(aes(fill=.estimate),lwd=0.1) +
  scale_fill_distiller(palette="Reds",direction=1,name="smape") +
  facet_wrap(vars(modelo), nrow = 1) +
  theme_linedraw(base_size = 10) +
  theme(strip.text = element_text(face = "bold",size = 20),
        legend.position = "bottom")
```



# cross validation ----
```{r}
cpo.10cp.sp <--2*sum(log(modelo_fh10_spatial$cpo$cpo),na.rm = T)
cpo.5cp.sp <--2*sum(log(modelo_fh5_spatial$cpo$cpo),na.rm = T)
cpo.8cp.sp <--2*sum(log(modelo_fh8_spatial$cpo$cpo),na.rm = T)
cpo.2cp.sp <--2*sum(log(modelo_fh2_spatial$cpo$cpo),na.rm = T)
cpo.2cp.sp.new <- -2*sum(log(modelo_fh2_spatial_new$cpo$cpo),na.rm = T)

data.cpo  <-  list(
                   "modelo5spat" = cpo.5cp.sp,
                   "modelo10spat" = cpo.10cp.sp,
                   "modelo8spat" = cpo.8cp.sp,
                   "modelo2spat" = cpo.2cp.sp,
                   "modelo2_new" = cpo.2cp.sp.new) %>% 
  as.data.frame()
```


```{r}
data.cpo %>% 
  pivot_longer(cols=colnames(data.cpo),
               names_to = "model",
               values_to = "CPO")  %>% 
  gt() %>%
  tab_header(title = md("LOO-CV")) %>% 
  tab_style(
    style = list(
      cell_fill(color = "#FFF3B0"),  # amarillo claro
      cell_text(weight = "bold")
    ),
    locations = cells_body(
      rows = model %in% c("modelo2spat")
    )
  )
```



## Visualization 
#### real values of edu level

```{r}
graficos<-
  distritos_sf_predict %>% 
  st_drop_geometry() %>% 
  select(year,ubigeo,real_cases,fit8_spat:fit2_new) %>% 
  pivot_longer(cols = fit8_spat:fit2_new) %>% 
  
  group_by(name) %>% 
  nest() %>% 
  
  mutate(
    data_graph = map(.x = data,
                     .f = ~.x %>% pivot_longer(cols = c(real_cases,value))),
    
    graph = map2(.x = data_graph, .y = name,
                .f = ~.x %>% ggplot(aes(x = value, y=ubigeo, col = name))+
                  geom_point(alpha = 0.5)+
                  geom_line(aes(group = ubigeo))+
                  facet_wrap(~year, scales = "free")+
                  ggtitle(name))
  )
```


```{r}
graficos$graph[[4]]
```

# calculation proportions with cases predicted

```{r}
prop_sae_predict_edu<-
  distritos_sf_predict %>% 
  select(year,ubigeo,pop_landscan,real_cases,fit10_spat,
         fit8_spat,fit5_spat,fit2_spat,fit2_new,geometry) %>% 
  mutate(
    across(.cols = c(real_cases:fit2_new), .f=~.x/pop_landscan)
  )
```


```{r}
prop_sae_predict_edu %>% 
  pivot_longer(cols = real_cases:fit2_new) %>% 
  mutate(
    name = factor(name, levels = c("real_cases", "fit2_new", "fit2_spatial", "fit5_spat", 
                                   "fit8_spat", "fit10_spat"))
  ) %>% 
  
  ggplot()+
  geom_sf(aes(fill = value)) +
  scale_fill_viridis(direction = -1,option = "rocket") +
  facet_grid(name~year)
```



# Point and interval estimates of proportions (INLA)

```{r}
modelos <- list( # lista para trabajar en bloque
  modelo_fh8_spatial = modelo_fh8_spatial,
  modelo_fh10_spatial = modelo_fh10_spatial,
  modelo_fh5_spatial = modelo_fh5_spatial,
  modelo_fh2_spatial = modelo_fh2_spatial,
  modelo_fh2_new = modelo_fh2_spatial_new
)
```


```{r}
resultados_SAE_loreto_edu <- 
  imap_dfr(modelos, ~{ # formato long de las estimaciones y los IC de cada modelo distrito-año
  data.frame(
    year = edu_censo_final$year,
    ubigeo = edu_censo_final$ubigeo,
    mean = .x$summary.fitted.values$mean,
    lower = .x$summary.fitted.values$`0.025quant`,
    upper = .x$summary.fitted.values$`0.975quant`,
    modelo = .y
  )
}) 

write.csv(resultados_SAE_loreto_edu, "./data/final_data/resultados_SAE_loreto_edu.csv", row.names = F )
```


```{r}
resultados_SAE_loreto_edu %>% 
  ggplot(aes(x = mean, y = ubigeo, col = modelo))+
  geom_point()+
  geom_errorbar(aes(xmin = lower, xmax = upper))+
  facet_grid(modelo~year)
```



# accuracy metris vs population
```{r}
tbl.yrd.dist %>% 
  full_join(
pop_dist %>% 
  group_by(ubigeo) %>% 
  summarise(pop_model = sum(pop_landscan))
) %>% 
  filter(
    modelo %in% c("pc2sp","pc2new","pc5sp","pc8sp","pc10sp")
  ) %>% 
  
  ggplot(aes(x = .estimate, y = log(pop_model), col = modelo))+
  geom_point(aes(shape = modelo), alpha = .5, size = 3)+
  geom_line(aes(group = ubigeo)) +
  
  coord_flip() +
  
  facet_wrap(~.metric, scales = "free", ncol = 1)
```

