---
title: "SAE: Education level"
output:
  html_document:
    toc: true
    toc_float: true
    number_sections: true
    theme: cosmo
date: "2025-05-28"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = F, message = F, fig.width = 12, fig.height = 8)
```


```{r}
library(tidyverse)
library(janitor)
library(sf)
library(innovar)
library(factoextra)
library(spdep)
library(INLA)
library(car)
library(yardstick)
library(gt)
library(viridis)
library(ggsci)
```

# Data
```{r}
data("Peru")
distritos<- Peru %>% filter(dep=="LORETO") %>% select(ubigeo,geometry)

aux_data<- read.csv("./data/aux_data/aux_data.csv") %>% mutate(ubigeo = as.character(ubigeo))

wi_edu_2010_2020<- read.csv("./data/wi_edu_2010_2020.csv")

wi_censo_final<- read.csv("./data/wi_censo_final_recat.csv") %>% mutate(ubigeo = as.character(ubigeo))

pop_dist <- read.csv("./data/aux_data/malaria_dist_total.csv") %>% 
  select(year,ubigeo,pop_landscan) %>% 
  mutate(
    ubigeo = as.character(ubigeo)
  )
```

# Work begins on the variable wealth index (hv270), recategorized as:belongs to 
# quintile 1 or 2 (1) and is in the 3rd or higher (0).
```{r, echo = F}
hist(wi_censo_final$hv270_recat) # no normal distribution
hist(wi_censo_final$hv270_recat_cor)  # no normal distribution

hist(wi_edu_2010_2020$hv271) # wi in continuous form. no normal distribution
```


Because neither has a normal distribution, even after transforming to the logit scale,
we explore other approaches (binomial, poisson, negative binomial).

For any of three options, we need:
  - numerator: cases
  - denominator: total population
 we have the proportion of cases (direct estimation from ENDES) and total population
from LandScan database.

To obtain the numerator: 
 ##proportion of cases * total population = estimated (hypothetical) number of cases

offset: log(total population)


# Fit INLA

## INLA - Education (working data)

```{r}
wi_censo_final2<-
  wi_censo_final %>% 
  left_join(aux_data) %>% ## adding auxiliary data
   inner_join(pop_dist) %>% ## Adding total population to the dataset (using malaria incidence data processed in the main repo)
  mutate(
    pop_landscan = round(pop_landscan),
    cases_number = round(hv270_recat*pop_landscan),
    year_re = as.integer(factor(year)), # for interaction spatial - time
    id.sp2 = id.sp # for interaction spatial - time
  )
```

## Spatial component

```{r}
loreto.dist <- poly2nb(distritos)
w.loreto <- nb2mat(loreto.dist, style = "W")
```


## Comparing family: binomial vs poisson vs negative binomial
```{r, echo = F}
# model_binom <- inla(
#   formula = cases_number ~ 1 + 
#     f(id.sp, model = "bym", graph = w.loreto) +   # efecto espacial
#     f(year, model = "ar1"),                       # efecto temporal
#   
#   data = wi_censo_final2,
#   family = "binomial",
#   Ntrials = pop_landscan,
#   
#   control.predictor = list(compute = TRUE, link = 1),
#   control.compute = list(dic = TRUE, waic = TRUE, cpo = TRUE)
# )
# 
# summary(model_binom)
# 
#  
# model_poisson <- inla(
#   formula = cases_number ~ 1 + 
#     f(id.sp, model = "bym", graph = w.loreto) +   # efecto espacial
#     f(year, model = "ar1"),                       # efecto temporal
#   
#   data = wi_censo_final2,
#   family = "poisson",
#   offset = log(pop_landscan),  
#   
#   control.predictor = list(compute = TRUE, link = 1),
#   control.compute = list(dic = TRUE, waic = TRUE, cpo = TRUE)
# )
# 
# summary(model_poisson)
# 
# model_nb <- inla(
#   formula = cases_number ~ 1 + 
#     f(id.sp, model = "bym2", graph = w.loreto) +
#     f(year_re, model = "ar1") +
#     f(id.sp2, model = "bym2", graph = w.loreto, group = year_re,
#       control.group = list(model = "ar1")),                       # efecto temporal
#   
#   data = wi_censo_final2,
#   family = "nbinomial",
#   offset = log(pop_landscan),  
#   
#   control.predictor = list(compute = TRUE, link = 1),
#   control.compute = list(dic = TRUE, waic = TRUE, cpo = TRUE)
# )
# 
# summary(model_nb) # The negative binomial family was chosen.
```


# Fit models (2) 
```{r}
model_fit10spat_1 <- inla(
  formula =  cases_number ~ 1 + pc1 + pc2 + pc3 + pc4 + pc5+ 
    pc6 + pc7 + pc8 + pc9 + pc10  +
    f(id.sp, model = "bym2", graph = w.loreto, group = year_re, # poner como menos 2000
      control.group = list(model = "ar1")),
  
  data = wi_censo_final2,
  family = "nbinomial",
  offset = log(pop_landscan),  
  
  control.predictor = list(compute = TRUE, link = 1),
  control.compute = list(dic = TRUE, cpo = TRUE, waic = TRUE, config = TRUE))
```


```{r}
model_fit10spat_2 <- inla(
  formula =  cases_number ~ 1 + pc1 + pc2 + pc3 + pc4 + pc5+ 
    pc6 + pc7 + pc8 + pc9 + pc10  +
    f(id.sp, model = "bym2", graph = w.loreto) +
    f(year_re, model = "ar1") +
    f(id.sp2, model = "bym2", graph = w.loreto, group = year_re,
      control.group = list(model = "ar1")),
  
  data = wi_censo_final2,
  family = "nbinomial",
  offset = log(pop_landscan),  
  
  control.predictor = list(compute = TRUE, link = 1),
  control.compute = list(dic = TRUE, cpo = TRUE, waic = TRUE, config = TRUE)
)

```


```{r}
model_fit8spat_1 <- inla(
  formula = cases_number ~ 1 + pc1 + pc2 + pc3 + pc4 + pc5 + 
    pc6 + pc7 + pc8 + total_millones + 
    f(id.sp, model = "bym2", graph = w.loreto, group = year_re,
      control.group = list(model = "ar1")),
  
  data = wi_censo_final2,
  family = "nbinomial",
  offset = log(pop_landscan),  
  
  control.predictor = list(compute = TRUE, link = 1),
  control.compute = list(dic = TRUE, waic = TRUE, cpo = TRUE,config = TRUE)
)

```


```{r}
model_fit8spat_2 <- inla(
  formula = cases_number ~ 1 + pc1 + pc2 + pc3 + pc4 + pc5 + 
    pc6 + pc7 + pc8 + total_millones + 
    f(id.sp, model = "bym2", graph = w.loreto) +
    f(year_re, model = "ar1") +
    f(id.sp2, model = "bym2", graph = w.loreto, group = year_re,
      control.group = list(model = "ar1")),
  
  data = wi_censo_final2,
  family = "nbinomial",
  offset = log(pop_landscan),  
  
  control.predictor = list(compute = TRUE, link = 1),
  control.compute = list(dic = TRUE, cpo = TRUE, waic = TRUE, config = TRUE)
)
```


```{r}
model_fit5spat_1 <- inla(
  formula = cases_number ~ 1 + pc1 + pc2 + pc3 + pc4 + pc5 + 
    f(id.sp, model = "bym2", graph = w.loreto, group = year_re,
      control.group = list(model = "ar1")),
  
  data = wi_censo_final2,
  family = "nbinomial",
  offset = log(pop_landscan),  
  
  control.predictor = list(compute = TRUE, link = 1),
  control.compute = list(dic = TRUE, waic = TRUE, cpo = TRUE,config = TRUE)
)

```


```{r}
model_fit5spat_2 <- inla(
  formula = cases_number ~ 1 + pc1 + pc2 + pc3 + pc4 + pc5 + 
    f(id.sp, model = "bym2", graph = w.loreto) +
    f(year_re, model = "ar1") +
    f(id.sp2, model = "bym2", graph = w.loreto, group = year_re,
      control.group = list(model = "ar1")),
  
  data = wi_censo_final2,
  family = "nbinomial",
  offset = log(pop_landscan),  
  
  control.predictor = list(compute = TRUE, link = 1),
  control.compute = list(dic = TRUE, cpo = TRUE, waic = TRUE, config = TRUE)
)
```


```{r}
model_fit2spat_1 <- inla(
  formula = cases_number ~ 1 + pc1 + pc2 +
    f(id.sp, model = "bym2", graph = w.loreto, group = year_re,
      control.group = list(model = "ar1")),
  
  data = wi_censo_final2,
  family = "nbinomial",
  offset = log(pop_landscan),
  
  control.predictor = list(compute = TRUE, link = 1),
  control.compute = list(dic = TRUE, waic = TRUE, cpo = TRUE, config = TRUE)
)

```


```{r}
model_fit2spat_2 <- inla(
  formula = cases_number ~ 1 + pc1 + pc2 + 
    f(id.sp, model = "bym2", graph = w.loreto) +
    f(year_re, model = "ar1") +
    f(id.sp2, model = "bym2", graph = w.loreto, group = year_re,
      control.group = list(model = "ar1")),
  
  data = wi_censo_final2,
  family = "nbinomial",
  offset = log(pop_landscan),  
  
  control.predictor = list(compute = TRUE, link = 1),
  control.compute = list(dic = TRUE, cpo = TRUE, waic = TRUE, config = TRUE)
)
```


```{r}
model_fit2spat_3 <- inla(
  formula =  cases_number ~ 1 + pc1 + pc2 + total_millones + edu_dist_mat +
    f(id.sp, model = "bym2", graph = w.loreto, group = year_re,
      control.group = list(model = "ar1")),
  
  data = wi_censo_final2,
  family = "nbinomial",
  offset = log(pop_landscan),  
  
  control.predictor = list(compute = TRUE, link = 1),
  control.compute = list(dic = TRUE, cpo = TRUE, waic = TRUE, config = TRUE)
)
```


```{r}
model_fit2spat_4<- inla(
  formula =  cases_number ~ 1 + pc1 + pc2 + total_millones + edu_dist_mat +
    f(id.sp, model = "bym2", graph = w.loreto) +
    f(year_re, model = "ar1") +
    f(id.sp2, model = "bym2", graph = w.loreto, group = year_re,
      control.group = list(model = "ar1")),
  
  data = wi_censo_final2,
  family = "nbinomial",
  offset = log(pop_landscan),  
  
  control.predictor = list(compute = TRUE, link = 1),
  control.compute = list(dic = TRUE, cpo = TRUE, waic = TRUE, config = TRUE)
)
```


# Summary

```{r}
model_metrics <- tibble(
    Modelo = c("Modelo 9", "Modelo 7",
             "Modelo 5", "Modelo 1",
             "Modelo 2", "Modelo 10",
             "Modelo 8","Modelo 6",
             "Modelo 3","Modelo 4"),
    
  DIC     = c(model_fit10spat_1$dic$dic, model_fit8spat_1$dic$dic, 
              model_fit5spat_1$dic$dic, model_fit2spat_1$dic$dic,model_fit2spat_2$dic$dic,
              model_fit10spat_2$dic$dic, model_fit8spat_2$dic$dic, 
              model_fit5spat_2$dic$dic, model_fit2spat_3$dic$dic,model_fit2spat_4$dic$dic),
  
  
  pD      = c(model_fit10spat_1$dic$p.eff, model_fit8spat_1$dic$p.eff, 
              model_fit5spat_1$dic$p.eff, model_fit2spat_1$dic$p.eff,model_fit2spat_2$dic$p.eff,
              model_fit10spat_2$dic$p.eff, model_fit8spat_2$dic$p.eff, 
              model_fit5spat_2$dic$p.eff, model_fit2spat_3$dic$p.eff,model_fit2spat_4$dic$p.eff),
  
  
  WAIC    = c(model_fit10spat_1$waic$waic, model_fit8spat_1$waic$waic, 
              model_fit5spat_1$waic$waic, model_fit2spat_1$waic$waic,model_fit2spat_2$waic$waic,
              model_fit10spat_2$waic$waic, model_fit8spat_2$waic$waic, 
              model_fit5spat_2$waic$waic, model_fit2spat_3$waic$waic,model_fit2spat_4$waic$waic),
  
  
  pWAIC   = c(model_fit10spat_1$waic$p.eff, model_fit8spat_1$waic$p.eff, 
              model_fit5spat_1$waic$p.eff, model_fit2spat_1$waic$p.eff,model_fit2spat_2$waic$p.eff,
              model_fit10spat_2$waic$p.eff, model_fit8spat_2$waic$p.eff, 
              model_fit5spat_2$waic$p.eff, model_fit2spat_3$waic$p.eff,model_fit2spat_4$waic$p.eff),
  
  
  logML   = c(model_fit10spat_1$mlik[1,1], model_fit8spat_1$mlik[1,1], 
              model_fit5spat_1$mlik[1,1], model_fit2spat_1$mlik[1,1],model_fit2spat_2$mlik[1,1],
              model_fit10spat_2$mlik[1,1], model_fit8spat_2$mlik[1,1], 
              model_fit5spat_2$mlik[1,1], model_fit2spat_3$mlik[1,1],model_fit2spat_4$mlik[1,1])
)
```

## Summary table
```{r}
model_metrics %>%
  arrange(WAIC) %>%
  gt() %>%
  tab_header(title = md("**Comparación de modelos INLA**")) %>%
  fmt_number(columns = c(DIC, pD, WAIC, pWAIC, logML), decimals = 2) %>%
  
  tab_style(
    style = list(
      # cell_fill(color = "#D5F5E3"),
      # cell_text(weight = "bold")
    ),
    locations = cells_body(
      #rows = Modelo %in% c("Modelo 8-2", "Modelo 10-2")
      )
  ) %>%
  
  tab_style(
    style = list(
      # cell_fill(color = "#FFF3B0"),
      # cell_text(weight = "bold")
    ),
    locations = cells_body(
      #rows = Modelo %in% c("Modelo 2-2", "Modelo 5-2","Modelo 10-1")
    )
  )
```


# Model accuracy (1: total values)
```{r}
distritos_sf_predict<-
  wi_censo_final2 %>% 
  mutate(
    fit10_spat = model_fit10spat_1$summary.fitted.values$mean,
    fit10_spat_2 = model_fit10spat_2$summary.fitted.values$mean,
    fit8_spat = (model_fit8spat_1$summary.fitted.values$mean),
    fit8_spat_2 = (model_fit8spat_2$summary.fitted.values$mean),
    fit5_spat = (model_fit5spat_1$summary.fitted.values$mean),
    fit5_spat_2 = (model_fit5spat_2$summary.fitted.values$mean),
    fit2_spat = (model_fit2spat_1$summary.fitted.values$mean),
    fit2_spat_2 = (model_fit2spat_2$summary.fitted.values$mean),
    fit2_spat_3 = (model_fit2spat_3$summary.fitted.values$mean),
    fit2_spat_4 = (model_fit2spat_4$summary.fitted.values$mean),
    real_cases = (wi_censo_final2$cases_number)
  ) %>% 
  left_join(distritos) %>% 
  st_as_sf()
```


```{r}
map.direct_est <-
  distritos_sf_predict %>% 
  filter(year == 2010) %>% 
  ggplot()+
  geom_sf(data = distritos)+
  geom_sf(aes(fill=cases_number))+
  scale_fill_viridis(direction = -1,option = "rocket") +
  theme(
    legend.position = "bottom"
  )
```


```{r}
map2sp <-
  distritos_sf_predict %>% 
  filter(year == 2010) %>% 
  ggplot()+
  geom_sf(data = distritos)+
  geom_sf(aes(fill=fit2_spat))+
  scale_fill_viridis(direction = -1,option = "rocket") +
  theme(
    legend.position = "bottom"
  )

map5sp <-
  distritos_sf_predict %>% 
  filter(year == 2010) %>% 
  ggplot()+
  geom_sf(data = distritos)+
  geom_sf(aes(fill=fit5_spat))+
  scale_fill_viridis(direction = -1,option = "rocket") +
  theme(
    legend.position = "bottom"
  )

map8sp <-
  distritos_sf_predict %>% 
  filter(year == 2010) %>% 
  ggplot()+
  geom_sf(data = distritos)+
  geom_sf(aes(fill=fit8_spat))+
  scale_fill_viridis(direction = -1,option = "rocket") +
  theme(
    legend.position = "bottom"
  )

map10sp <-
  distritos_sf_predict %>% 
  filter(year == 2010) %>% 
  ggplot()+
  geom_sf(data = distritos)+
  geom_sf(aes(fill=fit10_spat))+
  scale_fill_viridis(direction = -1,option = "rocket") +
  theme(
    legend.position = "bottom"
  )

map10sp_2 <-
  distritos_sf_predict %>% 
  filter(year == 2010) %>% 
  ggplot()+
  geom_sf(data = distritos)+
  geom_sf(aes(fill=fit10_spat_2))+
  scale_fill_viridis(direction = -1,option = "rocket") +
  theme(
    legend.position = "bottom"
  )
```


```{r}
cowplot::plot_grid(map.direct_est,map2sp,map5sp,map8sp,map10sp,map10sp_2)

#ggsave("./figures/prueba.png", dpi = 300, bg = "white")
```

## Comparison DE vs Pred. SAE ----
```{r}
distritos_sf_predict %>% 
  select(year,real_cases,fit10_spat,fit8_spat,fit5_spat,fit2_spat,fit2_spat_2) %>% 
  st_drop_geometry() %>%
  filter(!is.na(real_cases)) %>%
  rename(directo = real_cases) %>%
  pivot_longer(cols = starts_with("fit"),
               names_to = "modelo",
               values_to = "predicho") %>%
  ggplot(aes(x = directo, y = predicho, color = modelo)) +
  geom_point(alpha = 0.7, size = 2) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "gray40") +
  facet_grid(year~modelo) +
  labs(
    x = "Estimación directa",
    y = "Predicción del modelo",
    title = "Comparación de estimaciones directas vs modelos SAE",
    subtitle = "Línea diagonal = ajuste perfecto"
  ) +
  theme_minimal()
```


```{r}
distritos_sf_predict %>% 
  select(year,real_cases,fit10_spat_2,fit8_spat_2,fit5_spat_2,fit2_spat_4) %>% 
  st_drop_geometry() %>%
  filter(!is.na(real_cases)) %>%
  rename(directo = real_cases) %>%
  pivot_longer(cols = starts_with("fit"),
               names_to = "modelo",
               values_to = "predicho") %>%
  ggplot(aes(x = directo, y = predicho, color = modelo)) +
  geom_point(alpha = 0.7, size = 2) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "gray40") +
  facet_grid(year~modelo) +
  labs(
    x = "Estimación directa",
    y = "Predicción del modelo",
    title = "Comparación de estimaciones directas vs modelos SAE",
    subtitle = "Línea diagonal = ajuste perfecto"
  ) +
  theme_minimal()
```


## Performance metrics ----
```{r}

fitted_vals  <-  
distritos_sf_predict %>% 
  select(year,ubigeo,fit10_spat:real_cases) %>% 
  pivot_longer(cols = c(fit10_spat:fit2_spat_4), values_to = "fit", names_to = "modelo") %>% 
  mutate(
    actual = real_cases
  )
```


```{r}
#### Metrics ----
perform.metrics <- yardstick::metric_set(mae,mase,smape,rmse)
perform.metrics.dist <- yardstick::metric_set(mae,smape,rmse)

##### total metrics----
tbl.yrd.full <-  fitted_vals %>% 
  group_by(modelo) %>%
  perform.metrics(truth = actual, estimate = fit)

##### metrics by districts----
tbl.yrd.dist <-  fitted_vals %>% 
  group_by(modelo,ubigeo) %>%
  perform.metrics.dist(truth = actual, estimate = fit)
```



## Table in-sample accuracy metrics ----
```{r}
tbl.yrd.full %>% 
  pivot_wider(id_cols = modelo,
              names_from = .metric,
              values_from = .estimate) %>% 
  arrange(smape) %>% 
  gt() %>%
  tab_header(title = md("in-sample accuracy metrics")) %>% 
  tab_style(
    style = list(
      cell_fill(color = "#FFF3B0"),  # amarillo claro
      cell_text(weight = "bold")
    ),
    locations = cells_body(
      rows = modelo %in% c("pc2sp_4","pc2sp","pc5sp")
    )
  )
```

### mapping the performance metrics
```{r}
distritos %>% 
  inner_join(tbl.yrd.dist) %>% 
  filter(.metric=="smape") %>% 
  ggplot() +
  geom_sf(aes(fill=.estimate),lwd=0.1) +
  scale_fill_distiller(palette="Reds",direction=1,name="smape", limits = c(0,50)) +
  facet_wrap(vars(modelo), nrow = 2) +
  theme_linedraw(base_size = 10) +
  theme(
    strip.text = element_text(face = "bold",size = 30),
    legend.position = "bottom")
```

# cross validation
```{r}
cpo.10cp.sp <- -2*sum(log(model_fit10spat_1$cpo$cpo),na.rm = T)
cpo.10cp.sp2 <- -2*sum(log(model_fit10spat_2$cpo$cpo),na.rm = T)
cpo.8cp.sp <- -2*sum(log(model_fit8spat_1$cpo$cpo),na.rm = T)
cpo.8cp.sp2 <- -2*sum(log(model_fit8spat_2$cpo$cpo),na.rm = T)
cpo.5cp.sp <- -2*sum(log(model_fit5spat_1$cpo$cpo),na.rm = T)
cpo.5cp.sp2 <- -2*sum(log(model_fit5spat_2$cpo$cpo),na.rm = T)
cpo.2cp.sp <- -2*sum(log(model_fit2spat_1$cpo$cpo),na.rm = T)
cpo.2cp.sp2 <- -2*sum(log(model_fit2spat_2$cpo$cpo),na.rm = T)
cpo.2cp.sp3 <- -2*sum(log(model_fit2spat_3$cpo$cpo),na.rm = T)
cpo.2cp.sp4 <- -2*sum(log(model_fit2spat_4$cpo$cpo),na.rm = T)

data.cpo  <-  list("modelo9" = cpo.10cp.sp,
                   "modelo10" = cpo.10cp.sp2,
                   "modelo7" = cpo.8cp.sp,
                   "modelo8" = cpo.8cp.sp2,
                   "modelo5" = cpo.5cp.sp,
                   "modelo6" = cpo.5cp.sp2,
                   "modelo1" = cpo.2cp.sp,
                   "modelo2" = cpo.2cp.sp2,
                   "modelo3" = cpo.2cp.sp3,
                   "modelo4" = cpo.2cp.sp4) %>% 
  as.data.frame()
```


```{r}
data.cpo %>% 
  pivot_longer(cols=colnames(data.cpo),
               names_to = "model",
               values_to = "CPO")  %>% 
  arrange(CPO) %>% 
  gt() %>%
  tab_header(title = md("LOO-CV")) %>% 
  
  tab_style(
    style = list(
      cell_fill(color = "#FFF3B0"),  # amarillo claro
      cell_text(weight = "bold")
    ),
    locations = cells_body(
      rows = model %in% c("modelo2_2","modelo5_2","modelo10_2")
    )
  )

```

## Table of general metrics (1)
```{r}
bind_cols(
model_metrics %>% 
  arrange(Modelo),

tbl.yrd.full %>% 
  pivot_wider(id_cols = modelo,
              names_from = .metric,
              values_from = .estimate) %>% 
  arrange(modelo),

data.cpo %>% 
  pivot_longer(cols=colnames(data.cpo),
               names_to = "model",
               values_to = "CPO") %>% 
  arrange(model)
) %>% 
  select(Modelo,DIC,pD,WAIC,pWAIC,logML,mae,mase,smape,rmse,CPO) %>% 
  
  arrange(smape) %>% 
  gt() %>% 
  tab_header(title = md("**General metrics: model diagnosis, errors, LOO-CV**"),
             subtitle = md("**Count values**" )) %>% 
  tab_style(
    style = list(
       #cell_fill(color = "#FFF3B0"),  # amarillo claro
      # cell_text(weight = "bold")
    ),
    locations = cells_body(
      rows = Modelo %in% c("Modelo 10-2","Modelo 5-2","Modelo 2-2")
    )
  )
```

# Model accuracy (2: proportions values)

```{r}
distritos_sf_predict_prop<-
  wi_censo_final2 %>% 
  mutate(
    fit10_spat = model_fit10spat_1$summary.fitted.values$mean,
    fit10_spat_2 = model_fit10spat_2$summary.fitted.values$mean,
    fit8_spat = (model_fit8spat_1$summary.fitted.values$mean),
    fit8_spat_2 = (model_fit8spat_2$summary.fitted.values$mean),
    fit5_spat = (model_fit5spat_1$summary.fitted.values$mean),
    fit5_spat_2 = (model_fit5spat_2$summary.fitted.values$mean),
    fit2_spat = (model_fit2spat_1$summary.fitted.values$mean),
    fit2_spat_2 = (model_fit2spat_2$summary.fitted.values$mean),
    fit2_spat_3 = (model_fit2spat_3$summary.fitted.values$mean),
    fit2_spat_4 = (model_fit2spat_4$summary.fitted.values$mean),
    real_cases = (wi_censo_final2$cases_number),
    
    fit10_spatvar = model_fit10spat_1$summary.fitted.values$sd,
    fit10_spat_2var = model_fit10spat_2$summary.fitted.values$sd,
    fit8_spatvar = (model_fit8spat_1$summary.fitted.values$sd),
    fit8_spat_2var = (model_fit8spat_2$summary.fitted.values$sd),
    fit5_spatvar = (model_fit5spat_1$summary.fitted.values$sd),
    fit5_spat_2var = (model_fit5spat_2$summary.fitted.values$sd),
    fit2_spatvar = (model_fit2spat_1$summary.fitted.values$sd),
    fit2_spat_2var = (model_fit2spat_2$summary.fitted.values$sd),
    fit2_spat_3var = (model_fit2spat_3$summary.fitted.values$sd),
    fit2_spat_4var = (model_fit2spat_4$summary.fitted.values$sd),
    
  ) %>% 
  left_join(distritos) %>% 
  
  mutate(
    across(.cols = c(fit10_spat:real_cases), .f=~round(.x/pop_landscan,2)),
    across(.cols = c(fit10_spatvar:fit2_spat_4var), .f=~round((.x/pop_landscan)^2,6))
  ) %>% 
  st_as_sf()
```


```{r}
estimate_prop_wi <-
  distritos_sf_predict_prop %>% 
  select(ubigeo,year,real_cases,fit2_spat_2) %>% 
  filter(year %in% c(2010,2013,2017,2020)) %>% 
  pivot_longer(cols = c(real_cases:fit2_spat_2)) %>% 
  mutate(
    name = factor(name, levels = c("real_cases","fit2_spat_2"), labels = c("Direct estimate","Mp-2"))
  ) %>% 
  ggplot()+
  geom_sf(aes(fill=value))+
  scale_fill_gradient(low = "white", high = "#0061a9")+
  labs(fill = "Prop.") +
  theme_void()+
  theme(
    strip.text = element_text(face = "bold", size = 10)
  )+
  facet_grid(name~year)
```
### fig 6
```{r}
estimate_prop_edu
estimate_prop_wi

plot_grid(estimate_prop_edu,estimate_prop_wi,cols = 1, labels = c("A","B"))

ggsave("./figures/fig6.png", dpi = 800, bg = "white", width = 12,height = 10)
```

## Comparison DE vs Pred. SAE ----

```{r}
figura3_a<-
  distritos_sf_predict_prop %>% 
  select(year,real_cases,fit2_spat,fit2_spat_2,fit5_spat_2,fit8_spat_2,fit10_spat_2) %>% 
  st_drop_geometry() %>%
  filter(!is.na(real_cases)) %>%
  rename(directo = real_cases) %>%
  pivot_longer(cols = starts_with("fit"),
               names_to = "modelo",
               values_to = "predicho") %>%
  mutate(
    modelo = factor(modelo, 
                    levels = c("fit2_spat","fit2_spat_2","fit5_spat_2","fit8_spat_2","fit10_spat_2"),
                    labels = c("Mp-1","Mp-2","Mp-6","Mp-8","Mp-10"))
  ) %>% 
  ggplot(aes(x = directo, y = predicho, color = modelo)) +
  geom_point(alpha = 0.5, size = 2, show.legend = F) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "gray40") +
  scale_color_lancet() +
  facet_grid(year~modelo) +
  labs(
    x = "Prop.% (Direct estimation)",
    y = "Prop.% (Predicted)"
  ) +
  theme_minimal() +
  theme(legend.position = "bottom",
        axis.title = element_text(face = "bold"),
        axis.text = element_text(face = "bold"),
        strip.text = element_text(face ="bold"),
        legend.title = element_text(face = "bold"))
```

## Performance metrics ----

```{r}
fitted_vals  <-  
distritos_sf_predict_prop %>% 
  select(year,ubigeo,fit10_spat:real_cases) %>% 
  pivot_longer(cols = c(fit10_spat:fit2_spat_4), values_to = "fit", names_to = "modelo") %>% 
  mutate(
    actual = real_cases
  )
```


```{r}
#### Metrics ----
perform.metrics <- yardstick::metric_set(mae,mase,smape,rmse)
perform.metrics.dist <- yardstick::metric_set(mae,smape,rmse)
```


```{r}
##### total metrics----
tbl.yrd.full <-  fitted_vals %>% 
  group_by(modelo) %>%
  perform.metrics(truth = real_cases, estimate = fit)

##### metrics by districts----
tbl.yrd.dist_prop <-  fitted_vals %>% 
  group_by(modelo,ubigeo) %>%
  perform.metrics.dist(truth = real_cases, estimate = fit)

##### metrics by years----
tbl.yrd.full_year <-  fitted_vals %>% 
  group_by(modelo,year) %>%
  perform.metrics(truth = real_cases, estimate = fit)
```

## Table in-sample accuracy metrics ----

```{r}
tbl.yrd.full %>% 
  pivot_wider(id_cols = modelo,
              names_from = .metric,
              values_from = .estimate) %>%
  arrange(smape) %>% 
  gt() %>%
  tab_header(title = md("in-sample accuracy metrics")) %>% 
  tab_style(
    style = list(
      # cell_fill(color = "#FFF3B0"),  # amarillo claro
      # cell_text(weight = "bold")
    ),
    locations = cells_body(
      rows = modelo %in% c("fit8_spat_2","fit2_spat_2")
    )
  )
```

## mapping the performance metrics
```{r}
metrics_wi<-
  distritos %>% 
  inner_join(tbl.yrd.dist_prop) %>% 
  filter(modelo %in% c("fit2_spat","fit2_spat_2","fit5_spat_2","fit8_spat_2","fit10_spat_2")) %>%
  mutate(
    legend_title = .metric,
    modelo = factor(modelo, 
                    levels = c("fit2_spat","fit2_spat_2","fit5_spat_2","fit8_spat_2","fit10_spat_2"),
                    labels = c("Mp-1","Mp-2","Mp-6","Mp-8","Mp-10"))
  ) %>% 
  
  group_by(.metric) %>% 
  nest() %>% 
  mutate(
    grafico = map(.x =data,
                  .f =~ggplot(data = .x) +
                    geom_sf(aes(fill=.estimate)) +
                    scale_fill_gradient(low = "white", high = "#0061a9", name =.x$legend_title)+
                    facet_wrap(~modelo, nrow = 1) +
                    theme_void())
  )
  
metrics_wi_graph<-plot_grid(plotlist = metrics_wi$grafico, ncol = 1)
```

### fig 4
```{r}
metrics_edu_graph
metrics_wi_graph

plot_grid(metrics_edu_graph,metrics_wi_graph, ncol = 1, labels = c("A","B"))

ggsave("./figures/fig4.png", dpi = 800, width = 9,height = 14, bg = "white")

```

### heatplot accuracy metrics by year

```{r}
graph_heatplot<-
bind_rows(tbl.yrd.full_year,tbl.yrd.full) %>% 
  mutate(
    year2 = ifelse(is.na(year),"T. model",year)
  ) %>% 
  
  filter(
    modelo %in% c("fit2_spat","fit2_spat_2","fit5_spat_2","fit8_spat_2","fit10_spat_2") &
      
      .metric != "mase"
  ) %>% 
  mutate(
    labels = round(.estimate,3),
    legend_title = .metric,
    
    modelo = factor(modelo, 
                    levels = c("fit2_spat","fit2_spat_2","fit5_spat_2","fit8_spat_2","fit10_spat_2"),
                    labels = c("Mp-1","Mp-2","Mp-6","Mp-8","Mp-10"))
  ) %>% 
  group_by(.metric) %>% 
  nest() %>% 
  
  mutate(
    grafico = map(.x = data,
                  .f = ~ggplot(data = .x)+
                    geom_tile(aes(x = modelo, y = as.factor(year2), fill = .estimate), alpha = .8, show.legend = F) +
                    geom_text(aes(x = modelo, y = as.factor(year2),label =labels), size = 3) +
                    scale_fill_tw3("blue")+
                    labs(
                      fill = NULL,
                      y = NULL,
                      x = .x$legend_title
                    ) +
                    
                    theme_minimal() +
                    scale_x_discrete(position = "top")+
                    theme(
                      axis.title.x = element_text(face = "bold"),
                      axis.text = element_text(face = "bold", size = 8),
                      strip.text = element_text(face ="bold"),
                      legend.title = element_text(face = "bold", size = 8)
                    )
    )
  )

figura2_b<-cowplot::plot_grid(plotlist = graph_heatplot$grafico, rows = 1)
figura2_b
```

### Figure 3
```{r}
figure2_parte2<-
  cowplot::plot_grid(figura3_a,figura2_b, 
                   cols = 1, 
                   rel_heights = c(.65,.4))

figure2_parte1
figure2_parte2


plot_grid(figure2_parte1,figure2_parte2, ncol = 1, labels = c("A","B"))
ggsave("./figures/fig2.png", dpi = 800, width = 15, height = 17, bg = "white")
```

### mapping the performance metrics
```{r}
distritos %>% 
  inner_join(tbl.yrd.dist_prop) %>% 
  filter(.metric=="smape" & modelo %in% c("fit2_spat","fit2_spat_2","fit5_spat_2","fit8_spat_2","fit10_spat_2")) %>% 
  ggplot() +
  geom_sf(aes(fill=.estimate),lwd=0.1) +
  scale_fill_distiller(palette="Reds",direction=1,name="smape", limits = c(0,30)) +
  facet_wrap(vars(modelo), nrow = 1) +
  theme_minimal(base_size = 8) +
  theme(strip.text = element_text(face = "bold",size = 20),
        legend.position = "bottom")
```
## MSE (bias + variance)

```{r}
mse_area<-
  bind_cols(
distritos_sf_predict_prop %>% 
  select(year,ubigeo,fit10_spat:real_cases) %>% 
  pivot_longer(cols = c(fit10_spat:fit2_spat_4), values_to = "fit", names_to = "modelo"),

distritos_sf_predict_prop %>% 
  select(year,ubigeo,fit10_spatvar:fit2_spat_4var) %>% 
  
  pivot_longer(cols = c(fit10_spatvar:fit2_spat_4var), values_to = "fit_var", names_to = "modelo_var") %>% 
  select(-year,-ubigeo)
) %>% 
  select(year,ubigeo,real_cases,modelo,fit,fit_var) %>% 
  
  mutate(
    bias2 = (fit - real_cases)^2,
    variance = fit_var,
    mse = bias2 + variance
  )

mse_area_long<-
  mse_area %>% 
  pivot_longer(cols = c(mse,bias2, variance)) %>% 
   mutate(name = factor(name, levels = c("mse", "bias2", "variance"))) %>% 
  filter(
    modelo %in% c("fit2_spat","fit2_spat_2","fit5_spat_2","fit8_spat_2","fit10_spat_2")
  ) %>% 
  mutate(
    modelo = factor(modelo, 
                    levels = c("fit2_spat","fit2_spat_2","fit5_spat_2","fit8_spat_2","fit10_spat_2"),
                    labels = c("Mp-1","Mp-2","Mp-6","Mp-8","Mp-10"))
  )


mse_area_year<-
  mse_area %>% 
  group_by(modelo, year) %>%
  summarise(
    bias2 = mean(bias2, na.rm = TRUE),
    variance = mean(variance, na.rm = TRUE),
    mse = mean(mse, na.rm = TRUE),
    .groups = "drop"
  ) %>% 
  pivot_longer(cols = c(mse,bias2, variance)) %>% 
   mutate(name = factor(name, levels = c("mse", "bias2", "variance")),
          
          modelo = factor(modelo, 
                    levels = c("fit2_spat","fit2_spat_2","fit5_spat_2","fit8_spat_2","fit10_spat_2"),
                    labels = c("Mp-1","Mp-2","Mp-6","Mp-8","Mp-10"))) %>% 
  filter(
    modelo %in% c("Mp-1","Mp-2","Mp-6","Mp-8","Mp-10")
  )



labelsfacet2 <- c(
  "mse"     = "MSE",
  "bias2"   = "Bias^2",      # superíndice
  "variance"= "Variance"
)

mse_wi <-
  ggplot(mse_area_long, aes(x = as.character(year), y = log(value), 
                            color = modelo, shape = modelo)) +
  geom_point(alpha = 0.1) +
  geom_point(
    data = mse_area_year,
    aes(x = as.character(year), y = log(value)),
    size = 3,
    alpha = 0.9
  ) +
  scale_color_lancet() +
  labs(
    x = "years",
    y = "log(value)",
    color = "Model",
    shape = "Model"
  ) +
  facet_wrap(
    ~name, 
    labeller = labeller(name = labelsfacet2, .default = label_parsed)
  ) +
  theme_minimal() +
  theme(
    panel.grid.minor = element_blank(),
    axis.title = element_text(face = "bold"),
    strip.text = element_text(face = "bold", size = 12),
    axis.text = element_text(size = 8)
  )

mse_wi
```

### fig 5
```{r}
mse_wi
mse_edu

plot_grid(mse_edu,mse_wi, ncol = 1, labels = c("A","B"))

ggsave("./figures/fig5.png", dpi = 800, bg = "white", width = 12, height = 10)
```


# cross validation
```{r}
cpo.10cp.sp <- -2*sum(log(model_fit10spat_1$cpo$cpo),na.rm = T)
cpo.10cp.sp2 <- -2*sum(log(model_fit10spat_2$cpo$cpo),na.rm = T)
cpo.8cp.sp <- -2*sum(log(model_fit8spat_1$cpo$cpo),na.rm = T)
cpo.8cp.sp2 <- -2*sum(log(model_fit8spat_2$cpo$cpo),na.rm = T)
cpo.5cp.sp <- -2*sum(log(model_fit5spat_1$cpo$cpo),na.rm = T)
cpo.5cp.sp2 <- -2*sum(log(model_fit5spat_2$cpo$cpo),na.rm = T)
cpo.2cp.sp <- -2*sum(log(model_fit2spat_1$cpo$cpo),na.rm = T)
cpo.2cp.sp2 <- -2*sum(log(model_fit2spat_2$cpo$cpo),na.rm = T)
cpo.2cp.sp3 <- -2*sum(log(model_fit2spat_3$cpo$cpo),na.rm = T)
cpo.2cp.sp4 <- -2*sum(log(model_fit2spat_4$cpo$cpo),na.rm = T)

data.cpo  <-  list("modelo10" = cpo.10cp.sp,
                   "modelo10_2" = cpo.10cp.sp2,
                   "modelo8" = cpo.8cp.sp,
                   "modelo8_2" = cpo.8cp.sp2,
                   "modelo5" = cpo.5cp.sp,
                   "modelo5_2" = cpo.5cp.sp2,
                   "modelo2" = cpo.2cp.sp,
                   "modelo2_2" = cpo.2cp.sp2,
                   "modelo2_3" = cpo.2cp.sp3,
                   "modelo2_4" = cpo.2cp.sp4) %>% 
  as.data.frame()
```


```{r}
data.cpo %>% 
  pivot_longer(cols=colnames(data.cpo),
               names_to = "model",
               values_to = "CPO")  %>% 
  arrange(CPO) %>% 
  gt() %>%
  tab_header(title = md("LOO-CV")) %>% 
  tab_style(
    style = list(
      # cell_fill(color = "#FFF3B0"),  # amarillo claro
      # cell_text(weight = "bold")
    ),
    locations = cells_body(
      rows = model %in% c("modelo2spat_2")
    )
  )
```

## Table of general metrics (2)
```{r}
bind_cols(
model_metrics %>% 
  arrange(Modelo),

tbl.yrd.full %>% 
  pivot_wider(id_cols = modelo,
              names_from = .metric,
              values_from = .estimate) %>% 
  arrange(modelo),

data.cpo %>% 
  pivot_longer(cols=colnames(data.cpo),
               names_to = "model",
               values_to = "CPO") %>% 
  arrange(model)
) %>% 
  select(Modelo,DIC,pD,WAIC,pWAIC,logML,mae,mase,smape,rmse,CPO) %>% 
  
  arrange(smape) %>% 
  gt() %>% 
  tab_header(title = md("**General metrics: model diagnosis, errors, LOO-CV**"),
             subtitle = md("**Count values**" )) %>% 
  tab_style(
    style = list(
       #cell_fill(color = "#FFF3B0"),  # amarillo claro
      # cell_text(weight = "bold")
    ),
    locations = cells_body(
      rows = Modelo %in% c("Modelo 10-2","Modelo 5-2","Modelo 2-2")
    )
  )
```

# Visualization 

#### Real values of edu level

##### Absolute values
```{r}
# real values of edu level
graficos<-
  distritos_sf_predict %>% 
  st_drop_geometry() %>% 
  select(year,ubigeo,real_cases,fit10_spat,fit10_spat_2,fit8_spat,fit8_spat_2,
         fit5_spat,fit5_spat_2,fit2_spat,fit2_spat_2,fit2_spat_4) %>% 
  pivot_longer(cols = fit10_spat:fit2_spat_4) %>% 
  mutate(
    #ubigeo = as.numeric(ubigeo)
  ) %>% 
  
  group_by(name) %>% 
  nest() %>% 
  
  mutate(
    data_graph = map(.x = data,
                     .f = ~.x %>% pivot_longer(cols = c(real_cases,value))),
    
    graph = map2(.x = data_graph, .y = name,
                 .f = ~.x %>% ggplot(aes(x = value, y=ubigeo, col = name))+
                   geom_point(alpha = 0.5)+
                   geom_line(aes(group = ubigeo))+
                   facet_wrap(~year, scales = "free")+
                   ggtitle(name))
  )
```


```{r}
graficos$graph
```

##### Relative values (proportions)

```{r}
library(ggh4x)

graficos_prop<-
  distritos_sf_predict_prop %>% 
  st_drop_geometry() %>% 
  select(year,ubigeo,pop_landscan,real_cases,fit10_spat:fit2_spat_4) %>% 
  pivot_longer(cols = fit10_spat:fit2_spat_2) %>% 
  mutate(
    dif = real_cases - value
  ) %>% 
  
  group_by(name) %>% 
  nest() %>% 
  
  mutate(
    data_graph = map(.x = data,
                     .f = ~.x %>% pivot_longer(cols = c(real_cases,value))),
    
    graph = map2(.x = data_graph, .y = name,
                .f = ~.x %>% ggplot(aes(x = value, y=ubigeo, col = name))+
                  geom_point(alpha = 0.5, aes(size = pop_landscan))+
                  geom_line(aes(group = ubigeo))+
                  scale_color_discrete(
                    name = NULL,                # título
                    labels = c("estimacion_real" = "Direct estimate",
                               "value" = "Predicted value")) +
                  facet_wrap(~year, scales = "free",nrow = 2)+
                  scale_x_continuous(limits = c(0.2, 1), breaks = seq(0.2, 1, .1))+
                  facetted_pos_scales(
                    y = list(
                      `2010` = scale_y_discrete(),
                      `2011` = scale_y_discrete(labels = NULL, breaks = NULL),
                      `2012` = scale_y_discrete(labels = NULL, breaks = NULL),
                      `2013` = scale_y_discrete(labels = NULL, breaks = NULL),
                      `2014` = scale_y_discrete(labels = NULL, breaks = NULL),
                      `2015` = scale_y_discrete(labels = NULL, breaks = NULL),
                      `2016` = scale_y_discrete(),
                      `2017` = scale_y_discrete(labels = NULL, breaks = NULL),
                      `2018` = scale_y_discrete(labels = NULL, breaks = NULL),
                      `2019` = scale_y_discrete(labels = NULL, breaks = NULL),
                      `2020` = scale_y_discrete(labels = NULL, breaks = NULL))
                    ) +
                  theme_bw() +
                  theme(legend.position = "bottom",
                        axis.text.y = element_text(size = 7),
                        panel.grid = element_blank()))
  )
```

```{r, warning=F, message=F}
graficos_prop$graph[[8]]

ggsave("./figures/sm_fig7.png", dpi = 800, width = 17, height = 10)
```

#### Calculation proportions with cases predicted
```{r}
prop_sae_predict_wi<-
  distritos_sf_predict %>% 
  select(year,ubigeo,pop_landscan,real_cases,fit10_spat,fit10_spat_2,fit8_spat,fit8_spat_2,
         fit5_spat,fit5_spat_2,fit2_spat,fit2_spat_2,fit2_spat_4) %>% 
  mutate(
    across(.cols = c(real_cases:fit2_spat_4), .f=~.x/pop_landscan)
  )
```


```{r}
prop_sae_predict_wi %>%
  pivot_longer(cols = real_cases:fit2_spat_4) %>%
  mutate(
    name = factor(name, levels = c("real_cases", "fit10_spat",
                                   "fit10_spat_2", "fit8_spat",
                                   "fit8_spat_2", "fit5_spat",
                                   "fit5_spat_2","fit2_spat","fit2_spat_2","fit2_spat_4"))
  ) %>%

  ggplot()+
  geom_sf(aes(fill = value)) +
  scale_fill_viridis(direction = -1,option = "rocket") +
  facet_grid(name~year)
```


### Point and interval estimates of proportions (INLA)
```{r}
modelos <- list( # lista para trabajar en bloque
 
    modelo_fh10_spatial = model_fit10spat_1,
    modelo_fh10_spatial_2 = model_fit10spat_2,
    modelo_fh8_spatial = model_fit8spat_1,
    modelo_fh8_spatial_2 = model_fit8spat_2,
    modelo_fh5_spatial = model_fit5spat_1,
    modelo_fh5_spatial_2 = model_fit5spat_2,
    modelo_fh2_spatial = model_fit2spat_1,
    modelo_fh2_spatial_2 = model_fit2spat_2,
    modelo_fh2_spatial_4 = model_fit2spat_4
)
```


```{r}
resultados_SAE_loreto_wi <- 
  imap_dfr(modelos, ~{ # formato long de las estimaciones y los IC de cada modelo distrito-año
    data.frame(
      year = wi_censo_final2$year,
      ubigeo = wi_censo_final2$ubigeo,
      mean = .x$summary.fitted.values$mean,
      lower = .x$summary.fitted.values$`0.025quant`,
      upper = .x$summary.fitted.values$`0.975quant`,
      modelo = .y
    )
  }) %>% 
  inner_join(pop_dist) %>% 
  rename(
    wi_estimate = mean
  ) %>% 
  mutate(
    across(.cols = wi_estimate:upper, .f=~(.x/pop_landscan), .names = '{.col}_prop')
  )

#write.csv(resultados_SAE_loreto_wi, "./data/final_data/resultados_SAE_loreto_wi.csv", row.names = F )
```


```{r}
resultados_SAE_loreto_wi %>% 
  filter(modelo %in% c("modelo_fh2_spatial_2")) %>% 
  mutate(
    #wi_estimate_prop = pmin(pmax(wi_estimate_prop, 0), 1)
    #lower_prop = pmax(0, lower_prop),
    #upper_prop = pmin(1, upper_prop)
  ) %>% 
  
  full_join(wi_censo_final2 %>% 
  select(year,ubigeo,hv270_recat,se) %>% 
  mutate(
    low_real  = hv270_recat - (1.96 * se),
    high_real = hv270_recat + (1.96 * se)
  )) %>% 
  select(
    year,ubigeo,hv270_recat,low_real,high_real,wi_estimate_prop,lower_prop,upper_prop
  ) %>% 
  pivot_longer(
    cols = c(hv270_recat, wi_estimate_prop),
    names_to = "tipo",
    values_to = "estimate"
  ) %>%
  mutate(
    lower = case_when(
      tipo == "hv270_recat" ~ low_real,
      tipo == "wi_estimate_prop" ~ lower_prop
    ),
    upper = case_when(
      tipo == "hv270_recat" ~ high_real,
      tipo == "wi_estimate_prop" ~ upper_prop
    )
  ) %>% 

  ggplot(aes(x = estimate, y = ubigeo, col = tipo))+
  geom_pointrange(aes(xmin = lower, xmax = upper), alpha = .8) +
  scale_color_manual(
                    name = NULL,
                    values = c("hv270_recat" = "#75c9a6",
                               "wi_estimate_prop" = "#11464b"),
                    
                    labels = c("hv270_recat" = "Direct estimate",
                               "wi_estimate_prop" = "Predicted value")) +
  theme_bw()+
  facet_wrap(~year, ncol = 6)

ggsave("./figures/sm_fig9.png", dpi = 800, width = 17, height = 10)
```

# accuracy metris vs population

## count values
```{r}
tbl.yrd.dist %>% 
  full_join(
    pop_dist %>% 
      group_by(ubigeo) %>% 
      summarise(pop_model = sum(pop_landscan))
  ) %>% 
  filter(
    modelo %in% c("fit10_spat","fit5_spat",
                  "fit2_spat","fit2_spat_2","fit2_spat_4")
  ) %>% 
  
  ggplot(aes(x = .estimate, y = log(pop_model), col = modelo))+
  geom_point(aes(shape = modelo), alpha = .5, size = 3)+
  geom_line(aes(group = ubigeo)) +
  
  coord_flip() +
  
  facet_wrap(~.metric, scales = "free", ncol = 1)
```

## prop values
```{r}
errors_wi<-
  tbl.yrd.dist_prop %>% 
  full_join(
pop_dist %>% 
  group_by(ubigeo) %>% 
  summarise(pop_model = sum(pop_landscan))
) %>% 
  filter(
modelo %in% c("fit10_spat","fit5_spat",
                  "fit2_spat","fit2_spat_2","fit2_spat_4")
  ) %>% 
  
  ggplot(aes(x = .estimate, y = log(pop_model), col = modelo))+
  geom_point(aes(shape = modelo), alpha = .5, size = 3)+
  geom_line(aes(group = ubigeo)) +
  
  theme_bw()+
  
  scale_color_discrete(
    name = NULL,
    labels = c(
      "fit10_spat"   = "PC 10",
      "fit5_spat"    = "PC 5",
      "fit2_spat_4"  = "PC 2-4",
      "fit2_spat_2"  = "PC 2-2",
      "fit2_spat"    = "PC 2"
    )
  ) +
  
  scale_shape_discrete(
    name = NULL,
    labels = c(
      "fit10_spat"   = "PC 10",
      "fit5_spat"    = "PC 5",
      "fit2_spat_4"  = "PC 2-4",
      "fit2_spat_2"  = "PC 2-2",
      "fit2_spat"    = "PC 2"
    )
  ) +
  
  coord_flip() +
  
  facet_wrap(~.metric, scales = "free", ncol = 1)
```

```{r}
plot_grid(errors_edu,errors_wi, nrow = 2, labels = c("A","B"))

ggsave("./figures/sm_fig10.png", dpi = 800, bg = "white", height = 10, width = 10)
```

```{r}

```

