---
title: "Untitled"
output: html_document
date: "2025-05-30"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = F, message = F, fig.height= 10, fig.width= 8)
```


```{r}
library(tidyverse)
library(survey)
library(janitor)
library(sf)
library(viridis)
```

# data
```{r}
wi_edu_2010_2020<- read.csv("./data/wi_edu_2010_2020.csv")

resultados_SAE_loreto_edu <- read.csv("./data/final_data/resultados_SAE_loreto_edu.csv") %>% 
  mutate(ubigeo = as.character(ubigeo),
         ubigeo_prov = substr(ubigeo, 1, 4))



resultados_SAE_loreto_wi <- read.csv("./data/final_data/resultados_SAE_loreto_wi.csv") %>% 
  mutate(ubigeo = as.character(ubigeo),
         ubigeo_prov = substr(ubigeo, 1, 4))



pop_dist <- read.csv("./data/aux_data/malaria_dist_total.csv") %>% 
  select(year,iddist,pop_landscan) %>% 
  rename(ubigeo = iddist) %>% 
  mutate(
    ubigeo = as.character(ubigeo)
  )
```

# Aggregated consistency check - EDU

1. Estimates will be calculated for each year for the whole Loreto department and provinces, 
2. The district-level estimates by year will be aggregated to the department-year and province-year levels,
3. It will be verified if the direct estimate is on the confidence interval (CI) of the SAE-INLA model estimates.

The logic is that the true value is expected to lie within the CI of the model/design-based estimate.
ENDES assumes that the true value lies within its own CI. For the SAE-INLA models, we would expect 
ENDES estimates to fall within the model-generated CIs.

## Departament: Loreto
```{r}
# 1. Estimates will be calculated for each year for the whole Loreto department and provinces

data_final_nest<-
  wi_edu_2010_2020 %>% 
  group_by(year) %>% 
  nest() %>% 
  mutate(
    svydesign_endes= map(.x=data,
                         .f = ~svydesign(ids = ~hv001, strata = ~hv022, 
                                         weights = ~hv005, data = .x, nest = T)),
    
    wi_dep = map(.x = svydesign_endes,
                 .f = ~svyby(~hv270_recat, ~hv023, design = .x, svymean, keep.var = T)),
    
    edu_dep = map(.x = svydesign_endes,
                  .f = ~svyby(~hv109_recat, ~hv023, design = .x, svymean, keep.var = T)),
    
    
    wi_prov = map(.x = svydesign_endes,
                  .f = ~svyby(~hv270_recat, ~ubigeo_prov, design = .x, svymean, keep.var = T)),
    
    edu_prov = map(.x = svydesign_endes,
                  .f = ~svyby(~hv109_recat, ~ubigeo_prov, design = .x, svymean, keep.var = T))
    )
```


```{r}
# 2. The district-level estimates by year will be aggregated to the department-year level

predict_sae_loreto<-
  resultados_SAE_loreto_edu %>% 
  full_join(pop_dist) %>% # total population for calculate proportions
  group_by(year,modelo) %>% 
  summarise(
    total_loreto = sum(pop_landscan),
    mean_loreto = sum(mean),
    lower_loreto = sum(lower),
    upper_loreto = sum(upper)
  ) %>% 
  mutate(
    across(.cols = c(mean_loreto:upper_loreto), .f = ~.x/total_loreto)
  )
```


```{r}
#3. It will be verified if the direct estimate is on the confidence interval (CI) of 
# the SAE-INLA model estimates

validation_edu_depart<-
  
data_final_nest %>% 
  select(year,edu_dep) %>% 
  unnest(edu_dep) %>% 
  ungroup() %>%
  select(year,hv109_recat,se) %>% 
  
  full_join(predict_sae_loreto) %>% 
  
  mutate(
    check = ifelse(hv109_recat > lower_loreto & hv109_recat<upper_loreto, "cumple", "no cumple"),
    
    
    se_modelo = (upper_loreto - lower_loreto) / (2 * 1.96),
    z = (hv109_recat - mean_loreto) / sqrt(se^2 + se_modelo^2),
    p_value = 2 * pnorm(-abs(z)),
    
    diferencia_significativa = ifelse(p_value < 0.05,"S","NS")
    
  ) %>% 
  
  filter(
    modelo %in% c('modelo_fh10_spatial',
                  'modelo_fh8_spatial',
                  'modelo_fh5_spatial',
                  'modelo_fh2_spatial',
                  'modelo_fh2_new')
  )
```


```{r}
validation_edu_depart_graph<- # graph ----
  validation_edu_depart %>% 
  ggplot()+
  geom_pointrange(aes(x = modelo, y =mean_loreto,
                      col = modelo, ymin = lower_loreto, ymax = upper_loreto))+
  
  geom_pointrange(aes(x = modelo, y = hv109_recat, 
                      ymin = hv109_recat-(1.96*se), 
                      ymax = hv109_recat+(1.96*se),
                      shape = diferencia_significativa), 
                  
                  col = 'black', alpha = .5, show.legend = F)+
  
  scale_shape_manual(
    values = c(
      "S" = 23, 
      "NS" = 15)
  ) +
  
  coord_flip()+
  
  facet_grid(rows = vars(year), switch = "y") +
  
  theme_minimal() +
  
  theme(
    panel.grid.major=element_blank(),
    panel.grid.minor=element_blank(),
    axis.text.y=element_blank(),
    axis.line=element_line(),
    strip.placement = "outside",
    axis.text.x=element_text(face="bold"),
    strip.text.y = element_text(vjust = 1,size = 10,face="bold"),
    legend.position = "bottom")
```

# Aggregated consistency check - WI

## Departament: Loreto
```{r}
predict_sae_loreto<-
  resultados_SAE_loreto_wi %>% 
  full_join(pop_dist) %>% # total population for calculate proportions
  group_by(year,modelo) %>% 
  summarise(
    total_loreto = round(sum(pop_landscan)),
    mean_loreto = sum(mean),
    lower_loreto = sum(lower),
    upper_loreto = sum(upper)
  ) %>% 
  mutate(
    across(.cols = c(mean_loreto:upper_loreto), .f = ~.x/total_loreto)
  )
```


```{r}
validation_wi_dep<-
  data_final_nest %>% 
  select(year,wi_dep) %>% 
  unnest(wi_dep) %>% 
  ungroup() %>%
  select(year,hv270_recat,se) %>% 
  
  full_join(predict_sae_loreto) %>% 
  
  mutate(
    check = ifelse(hv270_recat > lower_loreto & hv270_recat<upper_loreto, "cumple", "no cumple"),
    across(.cols = c(hv270_recat,mean_loreto:upper_loreto), .f = ~round(.x,2)),
    
    se_modelo = (upper_loreto - lower_loreto) / (2 * 1.96),
    z = (hv270_recat - mean_loreto) / sqrt(se^2 + se_modelo^2),
    p_value = 2 * pnorm(-abs(z)),
    
    diferencia_significativa = ifelse(p_value < 0.05,"S","NS")
    
  ) %>% 
  
  filter(
    modelo %in% c('modelo_fh10_spatial',
                  'modelo_fh8_spatial',
                  'modelo_fh5_spatial',
                  'modelo_fh2_spatial',
                  'modelo_fh2_new')
  )
```


```{r}
validation_wi_dep_graph<- # graph ----
  validation_wi_dep %>% 
  ggplot()+
  geom_pointrange(aes(x = modelo, y =mean_loreto,
                      col = modelo, ymin = lower_loreto, ymax = upper_loreto))+
  
  geom_pointrange(aes(x = modelo, y = hv270_recat, 
                      ymin = hv270_recat-(1.96*se), 
                      ymax = hv270_recat+(1.96*se),
                      shape = diferencia_significativa), 
                  
                  col = 'black', alpha = .5)+
  
  scale_shape_manual(
    values = c(
      "S" = 23, 
      "NS" = 15)
  ) +
  
  guides(
    color = guide_legend(order = 1),
    shape = guide_legend(order = 2)
  ) +
  
  coord_flip()+
  
  
  facet_grid(rows = vars(year), switch = "y") +
  
  theme_minimal() +
  
  theme(
    panel.grid.major=element_blank(),
    panel.grid.minor=element_blank(),
    axis.text.y=element_blank(),
    axis.line=element_line(),
    strip.placement = "outside",
    axis.text.x=element_text(face="bold"),
    strip.text.y = element_text(vjust = 1,size = 10,face="bold"),
    legend.position = "bottom")
```



## Complete graph
```{r}
library(patchwork)

validation_wi_dep_graph
validation_edu_depart_graph

g1<-  
(validation_wi_dep_graph|validation_edu_depart_graph) + 
  plot_layout(guides = "collect") &
  theme(legend.position = "bottom",legend.box = "vertical")

g1

#ggsave("./figures/graph1.png", dpi = 300, bg = "white", height = 12, width = 10)
```

