---
title: "Untitled"
output: html_document
date: "2025-05-30"
---

```{r setup, include=FALSE}
#knitr::opts_chunk$set(echo = TRUE, warning = F, message = F, fig.height= 10, fig.width= 8)
```


```{r}
library(tidyverse)
library(survey)
library(janitor)
library(sf)
library(viridis)
library(ggsci)

source("./_functions/functions.R")
```

# data
```{r}
wi_edu_2010_2020<- read.csv("./data/wi_edu_2010_2020.csv")

pop_dist <- read.csv("./data/aux_data/malaria_dist_total.csv") %>% 
  select(year,ubigeo,pop_landscan) %>% 
  mutate(
    ubigeo = as.character(ubigeo)
  )
```

# Aggregated consistency check - EDU

1. Estimates will be calculated for each year for the whole Loreto department and provinces, 
2. The district-level estimates by year will be aggregated to the department-year and province-year levels,
3. It will be verified if the direct estimate is on the confidence interval (CI) of the SAE-INLA model estimates.

The logic is that the true value is expected to lie within the CI of the model/design-based estimate.
ENDES assumes that the true value lies within its own CI. For the SAE-INLA models, we would expect 
ENDES estimates to fall within the model-generated CIs.

## Departament: Loreto
```{r}
# 1. Estimates will be calculated for each year for the whole Loreto department and provinces

data_final_nest<-
  wi_edu_2010_2020 %>% 
  group_by(year) %>% 
  nest() %>% 
  mutate(
    svydesign_endes= map(.x=data,
                         .f = ~svydesign(ids = ~hv001, strata = ~hv022, 
                                         weights = ~hv005, data = .x, nest = T)),
    
    wi_dep = map(.x = svydesign_endes,
                 .f = ~svyby(~hv270_recat, ~hv023, design = .x, svymean, keep.var = T)),
    
    edu_dep = map(.x = svydesign_endes,
                  .f = ~svyby(~hv109_recat, ~hv023, design = .x, svymean, keep.var = T)),
    
    
    wi_prov = map(.x = svydesign_endes,
                  .f = ~svyby(~hv270_recat, ~ubigeo_prov, design = .x, svymean, keep.var = T)),
    
    edu_prov = map(.x = svydesign_endes,
                  .f = ~svyby(~hv109_recat, ~ubigeo_prov, design = .x, svymean, keep.var = T))
    )
```


```{r}
# 2. The district-level estimates by year will be aggregated to the department-year level

modelos<- list(model_pc2=modelo_fh2_spatial_edu_2,
               model_pc2_2=modelo_fh2_spatial_edu_4,
               model_pc5=modelo_fh5_spatial_edu_2,
               model_pc8=modelo_fh8_spatial_edu_2,
               model_pc10=modelo_fh10_spatial_edu_2)

predict_sae_loreto <- imap_dfr(
  .x = modelos,
  .f = ~ {
    resumen_anual_samples(modelo_inla = .x, data = edu_censo_final2) %>% 
      mutate(modelo = .y) 
  }
)
```


```{r}
#3. It will be verified if the direct estimate is on the confidence interval (CI) of 
# the SAE-INLA model estimates

validation_edu_depart<- 
  
data_final_nest %>% 
  select(year,edu_dep) %>% 
  unnest(edu_dep) %>% 
  ungroup() %>%
  select(year,hv109_recat,se) %>% 
  
  full_join(predict_sae_loreto) %>% 
  
  mutate(
    check = ifelse(hv109_recat > li_95 & hv109_recat<ls_95, "cumple", "no cumple"),
    
    
    se_modelo = (ls_95 - li_95) / (2 * 1.96),
    z = (hv109_recat - media) / sqrt(se^2 + se_modelo^2),
    p_value = 2 * pnorm(-abs(z)),
    
    diferencia_significativa = ifelse(p_value < 0.05,"S","NS")
    
  )
```

```{r, warning=F}
#### Metrics ----
perform.metrics <- yardstick::metric_set(mae,smape,rmse)
perform.metrics.dist <- yardstick::metric_set(mae,smape,rmse)

##### total metrics----
tbl.yrd.full <-  validation_edu_depart %>% 
  group_by(modelo,year) %>%
  perform.metrics(truth = hv109_recat, estimate = media)
```


```{r, warning=F}
labelsfacet <- c(
  "model_pc10" = "Model PC 10",
  "model_pc2" = "Model PC 02",
  "model_pc2_2" = "Model PC 02-2",
  "model_pc5" = "Model PC 05",
  "model_pc8" = "Model PC 08"
)

year_estimate_edu<-
tbl.yrd.full %>% 
  mutate(
    .estimate = round(.estimate,3)
  ) %>% 
  filter(.metric == "mae") %>% 
  
  full_join(
  validation_edu_depart, by = c("modelo","year")
  ) %>% 
  
  pivot_longer(cols = c(hv109_recat,media)) %>% 
  mutate(
mae2 = as.factor(ifelse(.estimate <=0.03,"< 0.3",
                                  ifelse(.estimate >0.03 & 
                                           .estimate <=0.05,"0.3-0.5",
                                         ifelse(.estimate >0.05 & 
                                                  .estimate <=0.07,"0.5-0.7", NA)))),
    
    
    modelo = factor(modelo, levels = c("model_pc2","model_pc2_2",
                                       "model_pc5","model_pc8",
                                       "model_pc10"))
  ) %>% 
  ggplot()+
  geom_point(aes(x = as.factor(year), y = value, col = name, size = mae2)) +
  # scale_y_continuous(limits = c(0.76,.82),
  #                    breaks = seq(0.76,.82, by = 0.02)) +
  scale_color_lancet() +
  theme_minimal() +
  labs(color = NULL,
       size = "MAE",
       x = "Year",
       y = "Prop.") +
  guides(
  color = guide_legend(order = 1),
  size = guide_legend(order = 2)
  ) +
  theme(
    legend.position = "bottom",
    panel.grid.minor= element_blank(),
    panel.grid.major.x = element_blank(),
    strip.text = element_text(face = "bold"),
    legend.title = element_text(face = "bold"),
    axis.title = element_text(face = "bold")
    #legend.box = "vertical"
  ) +
  facet_wrap(~modelo,scales = "free",  labeller = labeller(modelo = labelsfacet))
```

### graph edu
```{r}
validation_edu_depart_graph<-
  validation_edu_depart %>% 
  mutate(
    dif = hv109_recat - media
  ) %>% 
  
  ggplot()+
  geom_pointrange(alpha = 0.5, aes(x = modelo, y =media,
                      col = modelo, ymin = li_95, ymax = ls_95))+

  scale_y_continuous(limits = c(0.65,.85),
                     breaks = seq(0.65,.85, by = 0.02))+ 
  
  scale_color_manual(name = "model values",
                     values = c("model_pc2" = "gray",
                                "model_pc2_2" = "purple",
                                "model_pc5" = "red",
                                "model_pc8"="blue",
                                "model_pc10" = "green",
                                "hv270_recat" = "black")) +
  coord_flip()+
  
  facet_grid(rows = vars(year), switch = "y") +
  
  theme_minimal() +
  
  theme(
    panel.grid.major=element_blank(),
    panel.grid.minor=element_blank(),
    axis.text.y=element_blank(),
    axis.line=element_line(),
    strip.placement = "outside",
    axis.text.x=element_text(face="bold"),
    strip.text.y = element_text(vjust = 1,size = 10,face="bold"),
    panel.spacing = unit(1.2, "lines"),
    legend.position = "none")
```


```{r}
validation_edu_depart_graph2<-
  validation_edu_depart %>% 
  mutate(
    difference = as.factor(ifelse(hv109_recat - media <=0.03,"< 3%",
                                  ifelse(hv109_recat - media >0.03 & 
                                           hv109_recat - media <=0.05,"3-5%",
                                         ifelse(hv109_recat - media >0.05 & 
                                                  hv109_recat - media <=0.065,"5-7%", NA))))
  ) %>% 
  
  select(
    year,hv109_recat,media,modelo,difference
  ) %>% 
  pivot_longer(
    cols = hv109_recat:media
  ) %>% 
  mutate(
    name = ifelse(name == "media",modelo,name)
  ) %>% 
  
  ggplot(aes(x = modelo, y =value, color = name))+
  
  geom_point(alpha = .5, aes(size = difference))+
  geom_line(aes(group = modelo)) +

  
    scale_y_continuous(limits = c(0.65,.85),
                     breaks = seq(0.65,.85, by = 0.02)) +
  
  scale_size_manual(values = c("< 3%" = 1.5,
                               "3-5%"= 3,
                               "5-7%" = 5)) +
  
  scale_color_manual(name = "model values",
                     values = c("model_pc2" = "gray",
                                "model_pc2_2" = "purple",
                                "model_pc5" = "red",
                                "model_pc8"="blue",
                                "model_pc10" = "green",
                                "hv270_recat" = "black")) +

  
  coord_flip()+
  
  facet_grid(rows = vars(year), switch = "y") +
  
  theme_minimal() +
  
  theme(
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),
    axis.text.y = element_blank(),
    axis.line = element_line(),
    strip.placement = "outside",
    axis.text.x = element_text(face = "bold"),
    strip.text.y = element_text(vjust = 1, size = 10, face = "bold"),
    
    # # Color legend stays below
    legend.position = "bottom",
    legend.box = "vertical")
```

# Aggregated consistency check - WI

## Departament: Loreto
```{r}
modelos<- list(model_pc2=model_fit2spat_1,
               model_pc2_2=model_fit2spat_2,
               model_pc2_4=model_fit2spat_4,
               model_pc5=model_fit5spat_1,
               model_pc10=model_fit10spat_1)

predict_sae_loreto <- imap_dfr(
  .x = modelos,
  .f = ~ {
    resumen_anual_samples(modelo_inla = .x, data = wi_censo_final2) %>% 
      mutate(modelo = .y) 
  }
)
```


```{r}
validation_wi_dep<-
  data_final_nest %>% 
  select(year,wi_dep) %>% 
  unnest(wi_dep) %>% 
  ungroup() %>%
  select(year,hv270_recat,se) %>% 
  
  full_join(predict_sae_loreto) %>% 
  
  mutate(
    check = ifelse(hv270_recat > li_95 & hv270_recat<ls_95, "cumple", "no cumple"),
    
    
    se_modelo = (ls_95 - li_95) / (2 * 1.96),
    z = (hv270_recat - media) / sqrt(se^2 + se_modelo^2),
    p_value = 2 * pnorm(-abs(z)),
    
    diferencia_significativa = ifelse(p_value < 0.05,"S","NS")
    
  )
```

```{r, warning=F}
#### Metrics ----
perform.metrics <- yardstick::metric_set(mae,smape,rmse)
perform.metrics.dist <- yardstick::metric_set(mae,smape,rmse)

##### total metrics----
tbl.yrd.full <-  validation_wi_dep %>% 
  group_by(modelo,year) %>%
  perform.metrics(truth = hv270_recat, estimate = media)
```


```{r, warning=F}

labelsfacet2 <- c(
  "model_pc10" = "Model PC 10",
  "model_pc2" = "Model PC 02",
  "model_pc2_2" = "Model PC 02-2",
  "model_pc2_4" = "Model PC 02-4",
  "model_pc5" = "Model PC 05"
)

year_estimate_wi<-
tbl.yrd.full %>% 
  mutate(
    .estimate = round(.estimate,3)
  ) %>% 
  filter(.metric == "mae") %>% 
  
  full_join(
  validation_wi_dep, by = c("modelo","year")
  ) %>% 
  
  pivot_longer(cols = c(hv270_recat,media)) %>% 
  mutate(
    mae2 = as.factor(ifelse(.estimate <=0.03,"< 0.3",
                                  ifelse(.estimate >0.03 & 
                                           .estimate <=0.05,"0.3-0.5",
                                         ifelse(.estimate >0.05 & 
                                                  .estimate <=0.07,"0.5-0.7", NA)))),
    
    name = ifelse(name == "hv270_recat","Direct estimate","Predicted value"),
    
    modelo = factor(modelo, levels = c("model_pc2","model_pc2_2",
                                       "model_pc2_4","model_pc5",
                                       "model_pc10"))
  ) %>% 
  ggplot()+
  geom_point(aes(x = as.factor(year), y = value, col = name, size = mae2)) +
  # scale_y_continuous(limits = c(0.68,.85),
  #                    breaks = seq(0.68,.85, by = 0.02)) +
  scale_color_lancet() +
  theme_minimal() +
  labs(color = NULL,
       size = "MAE",
       x = "Year",
       y = "Prop.") +
  guides(
  color = guide_legend(order = 1),
  size = guide_legend(order = 2)
  ) +
  theme(
    legend.position = "bottom",
    panel.grid.minor= element_blank(),
    panel.grid.major.x = element_blank(),
    strip.text = element_text(face = "bold"),
    legend.title = element_text(face = "bold"),
    axis.title = element_text(face = "bold")
    #legend.box = "vertical"
  ) +
  facet_wrap(~modelo,scales = "free", labeller = labeller(modelo = labelsfacet2))


```

### graph wi
```{r}
validation_wi_dep_graph<- # graph ----
  validation_wi_dep %>% 
  mutate(
    dif = hv270_recat - media
  ) %>% 
  
  filter(
    modelo %in% c("model_pc10","model_pc8","model_pc5",
                  "model_pc2","model_pc2_2")
  ) %>% 
  
  ggplot()+
  geom_pointrange(alpha = 0.5, aes(x = modelo, y =media,
                      col = modelo, ymin = li_95, ymax = ls_95))+

  scale_y_continuous(limits = c(0.65,.85),
                     breaks = seq(0.65,.85, by = 0.02)) + 
  
  scale_color_manual(name = "model values",
                     values = c("model_pc2" = "gray",
                                "model_pc2_2" = "purple",
                                "model_pc5" = "red",
                                "model_pc8"="blue",
                                "model_pc10" = "green",
                                "hv270_recat" = "black")) +
  coord_flip()+
  
  facet_grid(rows = vars(year), switch = "y") +
  
  theme_minimal() +
  
  theme(
    panel.grid.major=element_blank(),
    panel.grid.minor=element_blank(),
    axis.text.y=element_blank(),
    axis.line=element_line(),
    strip.placement = "outside",
    axis.text.x=element_text(face="bold"),
    strip.text.y = element_text(vjust = 1,size = 10,face="bold"),
    panel.spacing = unit(1.2, "lines"),
    legend.position = "bottom",
    legend.box = "vertical")
```


```{r, warning=F}
validation_wi_dep_graph2<-
  validation_wi_dep %>% 
  
  filter(
    modelo %in% c("model_pc10","model_pc8","model_pc5",
                  "model_pc2","model_pc2_2")) %>% 
      
  mutate(
    difference = as.factor(ifelse(hv270_recat - media <=0.03,"< 3%",
                                  ifelse(hv270_recat - media >0.03 & 
                                           hv270_recat - media <=0.05,"3-5%",
                                         ifelse(hv270_recat - media >0.05 & 
                                                  hv270_recat - media <=0.065,"5-7%", NA))))
  ) %>% 
  
  select(
    year,hv270_recat,media,modelo,difference
  ) %>% 
  pivot_longer(
    cols = hv270_recat:media
  ) %>% 
  mutate(
    name = ifelse(name == "media",modelo,name)
  ) %>% 
  
  ggplot(aes(x = modelo, y =value, color = name))+
  
  geom_point(alpha = .5, aes(size = difference))+
  geom_line(aes(group = modelo)) +

  
  scale_y_continuous(limits = c(0.65,.85),
                     breaks = seq(0.65,.85, by = 0.02)) +
  
  scale_size_manual(values = c("< 3%" = 1.5,
                               "3-5%"= 3,
                               "5-7%" = 5)) +
  
  scale_color_manual(name = "model values",
                     values = c("model_pc2" = "gray",
                                "model_pc2_2" = "purple",
                                "model_pc5" = "red",
                                "model_pc8"="blue",
                                "model_pc10" = "green",
                                "hv270_recat" = "black")) +

  
  coord_flip()+
  
  facet_grid(rows = vars(year), switch = "y") +
  
  theme_minimal() +
  
  theme(
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),
    axis.text.y = element_blank(),
    axis.line = element_line(),
    strip.placement = "outside",
    axis.text.x = element_text(face = "bold"),
    strip.text.y = element_text(vjust = 1, size = 10, face = "bold"),
    
    # # Color legend stays below
    legend.position = "bottom",
    legend.box = "vertical"
    )

  
```


## Complete graph
```{r}
library(patchwork)
library(cowplot)

validation_edu_depart_graph
validation_edu_depart_graph2




g2 <- validation_edu_depart_graph + theme(legend.position = "none")
g3 <- validation_edu_depart_graph2 + theme(legend.position = "none")
g4 <- validation_wi_dep_graph + theme(legend.position = "none")
g5 <- validation_wi_dep_graph2 + theme(legend.position = "none")
legend <- get_legend2(validation_wi_dep_graph2) # get_legend function is broken. Use get_legend2 from stackoverflow. view: source("./_functions/functions.R")


plot_grid(plot_grid(g2,g3,g4,g5,nrow = 1,labels = c("A","","B","")),
          legend, nrow = 2, rel_heights =c(0.8,0.1))

#ggsave("./figures/graph1.png", dpi = 800, bg = "white", height = 10, width = 16)
```


```{r}
year_estimate_wi
year_estimate_edu

g1 <- year_estimate_wi + theme(legend.position = "none")
g2 <- year_estimate_edu + theme(legend.position = "none")
legend <- get_legend2(year_estimate_wi)

plot_grid(g2,g1,legend,ncol = 1, rel_heights = c(0.9,0.9,0.05), labels = c("A","B"))

#ggsave("./figures/fig7.png", dpi = 800, bg = "white",width = 14, height = 9)
```

```{r}

```

