---
title: "Untitled"
output: html_document
date: "2025-05-30"
---

```{r setup, include=FALSE}
#knitr::opts_chunk$set(echo = TRUE, warning = F, message = F, fig.height= 10, fig.width= 8)
```


```{r}
library(tidyverse)
library(survey)
library(janitor)
library(sf)
library(viridis)
library(ggsci)

source("./_functions/functions.R")
```

# data
```{r}
wi_edu_2010_2020<- read.csv("./data/wi_edu_2010_2020.csv")

pop_dist <- read.csv("./data/aux_data/malaria_dist_total.csv") %>% 
  select(year,ubigeo,pop_landscan) %>% 
  mutate(
    ubigeo = as.character(ubigeo)
  )
```

# Aggregated consistency check - EDU

1. Estimates will be calculated for each year for the whole Loreto department and provinces, 
2. The district-level estimates by year will be aggregated to the department-year and province-year levels,
3. It will be verified if the direct estimate is on the confidence interval (CI) of the SAE-INLA model estimates.

The logic is that the true value is expected to lie within the CI of the model/design-based estimate.
ENDES assumes that the true value lies within its own CI. For the SAE-INLA models, we would expect 
ENDES estimates to fall within the model-generated CIs.

## Departament: Loreto
```{r}
# 1. Estimates will be calculated for each year for the whole Loreto department and provinces

data_final_nest<-
  wi_edu_2010_2020 %>% 
  group_by(year) %>% 
  nest() %>% 
  mutate(
    svydesign_endes= map(.x=data,
                         .f = ~svydesign(ids = ~hv001, strata = ~hv022, 
                                         weights = ~hv005, data = .x, nest = T)),
    
    wi_dep = map(.x = svydesign_endes,
                 .f = ~svyby(~hv270_recat, ~hv023, design = .x, svymean, keep.var = T)),
    
    edu_dep = map(.x = svydesign_endes,
                  .f = ~svyby(~hv109_recat, ~hv023, design = .x, svymean, keep.var = T)),
    
    
    wi_prov = map(.x = svydesign_endes,
                  .f = ~svyby(~hv270_recat, ~ubigeo_prov, design = .x, svymean, keep.var = T)),
    
    edu_prov = map(.x = svydesign_endes,
                  .f = ~svyby(~hv109_recat, ~ubigeo_prov, design = .x, svymean, keep.var = T))
    )
```


```{r}
# 2. The district-level estimates by year will be aggregated to the department-year level

modelos<- list(modelo_fh2_spatial_edu_2=modelo_fh2_spatial_edu_2,
               modelo_fh2_spatial_edu_4=modelo_fh2_spatial_edu_4,
               modelo_fh5_spatial_edu_2=modelo_fh5_spatial_edu_2,
               modelo_fh8_spatial_edu_2=modelo_fh8_spatial_edu_2,
               modelo_fh10_spatial_edu_2=modelo_fh10_spatial_edu_2)

predict_sae_loreto <- imap_dfr(
  .x = modelos,
  .f = ~ {
    resumen_anual_samples(modelo_inla = .x, data = edu_censo_final2) %>% 
      mutate(modelo = .y) 
  }
)
```


```{r}
#3. It will be verified if the direct estimate is on the confidence interval (CI) of 
# the SAE-INLA model estimates

validation_edu_depart<- 
  
data_final_nest %>% 
  select(year,edu_dep) %>% 
  unnest(edu_dep) %>% 
  ungroup() %>%
  select(year,hv109_recat,se) %>% 
  
  full_join(predict_sae_loreto) %>% 
  
  mutate(
    check = ifelse(hv109_recat > li_95 & hv109_recat<ls_95, "cumple", "no cumple"),
    
    
    se_modelo = (ls_95 - li_95) / (2 * 1.96),
    z = (hv109_recat - media) / sqrt(se^2 + se_modelo^2),
    p_value = 2 * pnorm(-abs(z)),
    
    diferencia_significativa = ifelse(p_value < 0.05,"S","NS")
    
  )
```

```{r, warning=F}
#### Metrics ----
perform.metrics <- yardstick::metric_set(mae,smape,rmse)
perform.metrics.dist <- yardstick::metric_set(mae,smape,rmse)

##### total metrics----
tbl.yrd.full <-  validation_edu_depart %>% 
  group_by(modelo,year) %>%
  perform.metrics(truth = hv109_recat, estimate = media)
```


```{r, warning=F}
labelsfacet <- c(
  "modelo_fh2_spatial_edu_2" = "Me-2",
  "modelo_fh2_spatial_edu_4" = "Me-4",
  "modelo_fh5_spatial_edu_2" = "Me-5",
  "modelo_fh8_spatial_edu_2" = "Me-8",
  "modelo_fh10_spatial_edu_2" = "Me-10"
)
year_estimate_edu<-
  
tbl.yrd.full %>% 
  mutate(
    .estimate = round(.estimate,3)
  ) %>% 
  filter(.metric == "mae") %>% 
  
  full_join(
    validation_edu_depart %>% 
      mutate(
        low_direct = hv109_recat - 1.96*se,
        upp_direct = hv109_recat + 1.96*se
      ),
    
    by = c("modelo","year")
  ) %>% 
  
  select(year,modelo,hv109_recat,low_direct,upp_direct,
         media,li_95,ls_95,.estimate) %>% 
    mutate(
      mae2 = as.factor(ifelse(.estimate <=0.03,"<= 0.03",
                                  ifelse(.estimate >0.03 & 
                                           .estimate <=0.05,"0.03-0.05",
                                         ifelse(.estimate >0.05 & 
                                                  .estimate <=0.07,"0.05-0.07", NA)))),
    
    
    modelo = factor(modelo, levels = c("modelo_fh2_spatial_edu_2","modelo_fh2_spatial_edu_4",
                                       "modelo_fh5_spatial_edu_2","modelo_fh8_spatial_edu_2",
                                       "modelo_fh10_spatial_edu_2"))
  ) %>% 
  
  ggplot()+
  
  geom_pointrange(aes(x =hv109_recat, y = as.factor(year), 
                      xmin = low_direct, xmax = upp_direct,
                      color = "Direct estimate",
                      size = mae2),
                  alpha = .5, linewidth = 1.5) +
  geom_pointrange(aes(x =media, y = as.factor(year),
                      xmin = li_95, xmax = ls_95,
                      color = "Predicted value",
                      size = mae2),
                  alpha = .5) +
  
  scale_color_manual(values = c("Direct estimate" = "red4",
                                "Predicted value" = "blue4")) +
  
  scale_size_manual(
  values = c("<= 0.03" = 0.5,
             "0.03-0.05" = 0.8,
             "0.05-0.07" = 1.1),
  name = "MAE"
) +
  
  theme_minimal() +
  labs(color = NULL,
       x = "Year",
       y = "Prop.") +
  theme(
    legend.position = "bottom",
    panel.grid.minor= element_blank(),
    panel.grid.major.x = element_blank(),
    strip.text = element_text(face = "bold"),
    legend.title = element_text(face = "bold"),
    axis.title = element_text(face = "bold")
    #legend.box = "vertical"
  ) +
  facet_wrap(~modelo, scales = "free", labeller = labeller(modelo = labelsfacet)) +
  coord_flip()

```

# Aggregated consistency check - WI

## Departament: Loreto
```{r}
modelos2<- list(model_fit2spat_1=model_fit2spat_1,
               model_fit2spat_2=model_fit2spat_2,
               model_fit5spat_2=model_fit5spat_2,
               model_fit8spat_2=model_fit8spat_2,
               model_fit10spat_2=model_fit10spat_2)

predict_sae_loreto <- imap_dfr(
  .x = modelos2,
  .f = ~ {
    resumen_anual_samples(modelo_inla = .x, data = wi_censo_final2) %>% 
      mutate(modelo = .y) 
  }
)
```


```{r}
validation_wi_dep<-
  data_final_nest %>% 
  select(year,wi_dep) %>% 
  unnest(wi_dep) %>% 
  ungroup() %>%
  select(year,hv270_recat,se) %>% 
  
  full_join(predict_sae_loreto) %>% 
  
  mutate(
    check = ifelse(hv270_recat > li_95 & hv270_recat<ls_95, "cumple", "no cumple"),
    
    
    se_modelo = (ls_95 - li_95) / (2 * 1.96),
    z = (hv270_recat - media) / sqrt(se^2 + se_modelo^2),
    p_value = 2 * pnorm(-abs(z)),
    
    diferencia_significativa = ifelse(p_value < 0.05,"S","NS")
    
  )
```

```{r, warning=F}
#### Metrics ----
perform.metrics <- yardstick::metric_set(mae,smape,rmse)
perform.metrics.dist <- yardstick::metric_set(mae,smape,rmse)

##### total metrics----
tbl.yrd.full <-  validation_wi_dep %>% 
  group_by(modelo,year) %>%
  perform.metrics(truth = hv270_recat, estimate = media)
```


```{r, warning=F}

labelsfacet2 <- c(
  "model_fit2spat_1" = "Mp-1",
  "model_fit2spat_2" = "Mp-2",
  "model_fit5spat_2" = "Mp-6",
  "model_fit8spat_2" = "Mp-8",
  "model_fit10spat_2" = "Mp-10"
)

year_estimate_wi<-
tbl.yrd.full %>% 
  mutate(
    .estimate = round(.estimate,3)
  ) %>% 
  filter(.metric == "mae") %>% 
  
  full_join(
    validation_wi_dep %>% 
      mutate(
        low_direct = hv270_recat - 1.96*se,
        upp_direct = hv270_recat + 1.96*se
      ),
    
    by = c("modelo","year")
  ) %>% 
  
  select(year,modelo,hv270_recat,low_direct,upp_direct,
         media,li_95,ls_95,.estimate) %>% 
    mutate(
      mae2 = as.factor(ifelse(.estimate <=0.03,"<= 0.03",
                                  ifelse(.estimate >0.03 & 
                                           .estimate <=0.05,"0.03-0.05",
                                         ifelse(.estimate >0.05 & 
                                                  .estimate <=0.07,"0.05-0.07", NA)))),
    
    
    modelo = factor(modelo, levels = c("model_fit2spat_1","model_fit2spat_2",
                                       "model_fit5spat_2","model_fit8spat_2",
                                       "model_fit10spat_2"))
  ) %>% 
  
  ggplot()+
  
  geom_pointrange(aes(x =hv270_recat, y = as.factor(year), 
                      xmin = low_direct, xmax = upp_direct,
                      color = "Direct estimate",
                      size = mae2),
                  alpha = .5, linewidth = 1.5) +
  geom_pointrange(aes(x =media, y = as.factor(year),
                      xmin = li_95, xmax = ls_95,
                      color = "Predicted value",
                      size = mae2),
                  alpha = .5) +
  
  scale_color_manual(values = c("Direct estimate" = "red4",
                                "Predicted value" = "blue4")) +
  
  scale_size_manual(
  values = c("<= 0.03" = 0.5,
             "0.03-0.05" = 0.8,
             "0.05-0.07" = 1.1),
  name = "MAE"
) +
  
  theme_minimal() +
  labs(color = NULL,
       x = "Year",
       y = "Prop.") +
  theme(
    legend.position = "bottom",
    panel.grid.minor= element_blank(),
    panel.grid.major.x = element_blank(),
    strip.text = element_text(face = "bold"),
    legend.title = element_text(face = "bold"),
    axis.title = element_text(face = "bold")
    #legend.box = "vertical"
  ) +
  facet_wrap(~modelo, scales = "free", labeller = labeller(modelo = labelsfacet2)) +
  coord_flip()
```

```{r}
year_estimate_wi
year_estimate_edu

g1 <- year_estimate_wi + theme(legend.position = "none")
g2 <- year_estimate_edu + theme(legend.position = "none")
legend <- get_legend2(year_estimate_wi)

plot_grid(g2,g1,legend,ncol = 1, rel_heights = c(0.9,0.9,0.05), labels = c("A","B"))

ggsave("./figures/fig7.png", dpi = 800, bg = "white",width = 14, height = 9)
```

```{r}

```

